---
title: "mrIML Case Study 2 - Forest Communities"
subtitle: "Hierarchical Model"
author: "Simon Riley, Ryan Leadbetter, Nick Fountain-Jones"
date: "2026-03-11"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r libraries, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Data Reading

# General Libraries
library(future)
library(tidyverse)

# Data Manipulation
#library(dplyr) # In tidymodels
library(sf)

# Modelling
library(tidymodels)
library(mrIML)
# Visualisation
library(ggrepel)
library(ggplot2)

## User parameters
bsNum = 5


# Set up parallel processing
n_cores <- parallel::detectCores()
plan("multisession", workers = n_cores-2)

timings <- list()
timingsRel <- list()

timings$start <- Sys.time()
timingsRel$start <- Sys.time()
```

This case study demonstrates *MrIML 2.0* on a larger dataset with
multiple taxonomic groups across a forest successional gradient. We
analyze 144 plots established across forest age categories:

-   **Mature forest**: Unharvested with no fire disturbance for ≥60
    years
-   **\~45 years post-harvest**
-   **\~27 years post-harvest**
-   **\~7 years post-harvest**

At each plot, we recorded ground-active beetles, plants, and birds,
along with environmental co-variates including coarse wood debris, leaf
litter depth, and soil variables (pH, conductivity).

## Data Preparation

```{r data import, echo = FALSE}
currentDir <- getwd()
loadFile <- paste0(currentDir,"/data/XYForrestDataA.rdata")
load(loadFile)

```

For beetle and bird environmental variables we remove all the beetle and
bird related columns from the predictors.For the plant environmental
variables all plant or moss related columns are removed.

```{r sorting variables, echo = TRUE}
X_beetles <- data_site  %>%
  dplyr::select(
    # -Beetles_S_Mat_Indval2,
    # -Beetle_S_Total,
    # -dist_regrowth,
    -Beetles_Mat_Ratio,
    # -Birds_S_Mat_EFL,
    # -Birds_S_Total,
    -Birds_Mat_Ratio,
    -Slope_F,
    -Aspect_Cor)


X_birds <- X_beetles

X_plants <- data_site %>%
  select(
    -contains("Bryo"),
    -contains("Moss"),
    -contains('Plant'),
    -Vegetation_cover)

```

For the occurrence data we remove the columns that are either so
prevalent they would likely always be present or so rare that they would
be absent.

```{r setting up occurance data, echo = TRUE}

Y_beetles <- data_beetles %>%
  dplyr::select(-Site, -Distance, -Transect, -Age) %>%
  dplyr::mutate_all(~ ifelse(. > 0, 1, .)) %>%
  mrIML::filterRareCommon(lower = 0.2, higher = 0.7)


Y_birds <- data_birds %>%
  dplyr::select(-Site, -Distance, -Transect, -Age) %>%
  mutate_all(~ ifelse(. > 0, 1, .)) %>%
  mrIML::filterRareCommon(lower = 0.2, higher = 0.7)

Y_plants <- data_plants %>%
  dplyr::select(-Site, -Distance, -Transect, -Age) %>%
  mutate_all(~ ifelse(. > 0, 1, .))%>%
  mrIML::filterRareCommon(lower = 0.2, higher = 0.7)

X1_beetles <- Y_beetles
X1_birds <- Y_birds
X1_plants <- Y_plants

#Put Boruta here if required?

```

```{r Model Setup}

timings$modelStart <- Sys.time()
timingsRel$modelStart <- Sys.time()-timingsRel$Start
## build generic random forest

model_rf <- rand_forest(
  trees = 1000,
  mode = "classification",
  mtry = tune(),
  min_n = tune()
) %>%
  set_engine("randomForest")
```

```{r Application of Models, echo = TRUE, results='hide'}


# Applying generic model
## Building beetles models

# Model 1: Environment only
yhats_beetles_env <- mrIMLpredicts(
  X = X_beetles,
  Y = Y_beetles,
  Model = model_rf,
  balance_data = 'no',
  tune_grid_size = 5,
  prop = 0.6,
  k = 5,
  racing = TRUE
)

# Model 2: Co-occurrence only
yhats_beetles_cooccur <- mrIMLpredicts(
  Y = Y_beetles,
  X1 = X1_beetles,
  Model = model_rf,
  balance_data = 'no',
  tune_grid_size = 5,
  prop = 0.6,
  k = 5,
  racing = TRUE
)

# Model 3: Combined
yhats_beetles_combined <- mrIMLpredicts(
  X = X_beetles,
  Y = Y_beetles,
  X1 = X1_beetles,
  Model = model_rf,
  balance_data = 'no',
  tune_grid_size = 5,
  prop = 0.6,
  k = 5,
  racing = TRUE
)

## Building bird models

# Model 1: Environment only
yhats_birds_env <- mrIMLpredicts(
  X = X_birds,
  Y = Y_birds,
  Model = model_rf,
  balance_data = 'no',
  tune_grid_size = 5,
  prop = 0.6,
  k = 5,
  racing = TRUE
)

# Model 2: Co-occurrence only
yhats_birds_cooccur <- mrIMLpredicts(
  Y = Y_birds,
  X1 = X1_birds,
  Model = model_rf,
  balance_data = 'no',
  tune_grid_size = 5,
  prop = 0.6,
  k = 5,
  racing = TRUE
)

# Model 3: Combined
yhats_birds_combined <- mrIMLpredicts(
  X = X_birds,
  Y = Y_birds,
  X1 = X1_birds,
  Model = model_rf,
  balance_data = 'no',
  tune_grid_size = 5,
  prop = 0.6,
  k = 5,
  racing = TRUE
)


# Train models
yhats_plants_env <- mrIMLpredicts(
  X = X_plants,
  Y = Y_plants,
  Model = model_rf,
  balance_data = 'no',
  tune_grid_size = 5,
  prop = 0.6,
  k = 5,
  racing = TRUE
)

yhats_plants_cooccur <- mrIMLpredicts(
  Y = Y_plants,
  X1 = X1_plants,
  Model = model_rf,
  balance_data = 'no',
  tune_grid_size = 5,
  prop = 0.6,
  k = 5,
  racing = TRUE
)

yhats_plants_combined <- mrIMLpredicts(
  X = X_plants,
  Y = Y_plants,
  X1 = X1_plants,
  Model = model_rf,
  balance_data = 'no',
  tune_grid_size = 5,
  prop = 0.6,
  k = 5,
  racing = TRUE
)





timings$modelsTrained = Sys.time()

timings$modelsTrained <- Sys.time()-timingsRel$modelStart
```

To assess the uncertainty bootstrapping is implemeneted.

```{r bootstrap, echo = TRUE}


bs_beetles <- mrBootstrap(yhats_beetles_combined, num_bootstrap = bsNum)
bs_plants <- mrBootstrap(yhats_plants_combined, num_bootstrap = bsNum)
bs_birds <- mrBootstrap(yhats_birds_combined, num_bootstrap = bsNum)

timings$bootstrap = Sys.time()

```

Once the processing has been completed we can evaluate the performance
of the different models

```{r Performance evaluation, echo = TRUE}

# Evaluate model performance beetles
ModelPerf_beetles_env <- mrIMLperformance(yhats_beetles_env)
ModelPerf_beetles_cooccur <- mrIMLperformance(yhats_beetles_cooccur)
ModelPerf_beetles_combined <- mrIMLperformance(yhats_beetles_combined)


# Performance comparison Beetles
perf_comparison_beetles <- mrPerformancePlot(
  ModelPerf1 = ModelPerf_beetles_combined,
  ModelPerf2 = ModelPerf_beetles_env,
  mode = "classification"
)


# Summary table
beetle_performance <- data.frame(
  Model = c("Environment only", "Co-occurrence only", "Combined"),
  MCC = c(
    ModelPerf_beetles_env[[2]],
    ModelPerf_beetles_cooccur[[2]],
    ModelPerf_beetles_combined[[2]]
  )
)


# Performance assessment
ModelPerf_birds_env <- mrIMLperformance(yhats_birds_env)
ModelPerf_birds_cooccur <- mrIMLperformance(yhats_birds_cooccur)
ModelPerf_birds_combined <- mrIMLperformance(yhats_birds_combined)




bird_performance <- data.frame(
  Model = c("Environment only", "Co-occurrence only", "Combined"),
  MCC = c(
    ModelPerf_birds_env[[2]],
    ModelPerf_birds_cooccur[[2]],
    ModelPerf_birds_combined[[2]]
  )
)

# Performance assessment
ModelPerf_plants_env <- mrIMLperformance(yhats_plants_env)
ModelPerf_plants_cooccur <- mrIMLperformance(yhats_plants_cooccur)
ModelPerf_plants_combined <- mrIMLperformance(yhats_plants_combined)

plant_performance <- data.frame(
  Model = c("Environment only", "Co-occurrence only", "Combined"),
  MCC = c(
    ModelPerf_plants_env[[2]],
    ModelPerf_plants_cooccur[[2]],
    ModelPerf_plants_combined[[2]]
  )
)

timings$performance <- Sys.time()
```

Variable importance is assessed by the following method
\<<This section needs
improvement>\>

```{r Variable Importance, echo = TRUE}


# Variable importance
vi_beetles <- mrVip(
  yhats_beetles_combined,
  mrBootstrap_obj = bs_beetles,
  threshold = 0.5,
  global_top_var = 10
)

vi_plants <- mrVip(
  yhats_plants_combined,
  mrBootstrap_obj = bs_plants,
  threshold = 0.5,
  global_top_var = 10
)


timings$vip <- Sys.time()

```

Finding co-occurance networks

```{r Co-occurance, echo = TRUE}

# Co-occurrence network
assoc_net_beetles <- mrCoOccurNet(bs_beetles)
assoc_net_beetles_filtered <- assoc_net_beetles %>%
  filter(mean_strength > 0.1)

assoc_net_plants <- mrCoOccurNet(bs_plants)
assoc_net_plants_filtered <- assoc_net_plants %>%
  filter(mean_strength > 0.1)

# Analyze interactions in plant community
int_plants <- mrInteractions(
  yhats_plants_combined,
  num_bootstrap = bsNum,
  feature = 'Nothcunn',
  top_int = 10
)

int_plants[[1]] /
  int_plants[[2]] /
  int_plants[[3]] 
# +
  # plot_annotation(title = "Plant Community Interactions")


```

```{r temporary home for tables}
cat("Beetle Performance:\n")
beetle_performance
vi_beetles[[3]]

cat("Bird Community Performance:\n")
print(bird_performance)
vi_plants[[3]] +
  labs(title = "Plant Community Variable Importance")

cat("Plant Community Performance:\n")
print(plant_performance)

# Combine performance results
performance_summary <- data.frame(
  Community = rep(c("Beetles", "Birds", "Plants"), each = 3),
  Model = rep(c("Environment only", "Co-occurrence only", "Combined"), 3),
  MCC = c(
    beetle_performance$MCC,
    bird_performance$MCC,
    plant_performance$MCC
  )
)

# Calculate improvement from adding co-occurrence
improvement_summary <- performance_summary %>%
  pivot_wider(names_from = Model, values_from = MCC) %>%
  mutate(
    Cooccur_improvement = `Co-occurrence only` - `Environment only`,
    Combined_improvement = Combined - `Environment only`
  )

# Create summary table
results_table <- data.frame(
  Taxonomic_Group = c("Beetles", "Birds", "Plants"),
  Best_Model_MCC = c(
    max(beetle_performance$MCC),
    max(bird_performance$MCC),
    max(plant_performance$MCC)
  ),
  Cooccurrence_Effect = c(
    improvement_summary$Combined_improvement[1],
    improvement_summary$Combined_improvement[2],
    improvement_summary$Combined_improvement[3]
  ),
  Primary_Drivers = c(
    "Soil characteristics, microclimate",
    "Altitude, unmeasured structural conditions",
    "Forest age, succession gradient"
  )
)


print(improvement_summary)




```

```{r temporary home for plots}

perf_comparison_beetles$performance_plot +
  labs(title = "Beetle Community Model Performance")
# 
# ## Network visualization Beetles
# if (nrow(assoc_net_beetles_filtered) > 0) {
#   g_beetles <- graph_from_data_frame(
#     assoc_net_beetles_filtered,
#     directed = TRUE
#   )
# 
#   E(g_beetles)$Value <- assoc_net_beetles_filtered$mean_strength
#   E(g_beetles)$Color <- ifelse(
#     assoc_net_beetles_filtered$direction == "negative",
#     no = "blue",
#     yes = "red"
#   )
# 
#   gg_beetles <- ggnetwork(g_beetles)
# 
#   ggplot(gg_beetles, aes(x = x, y = y, xend = xend, yend = yend)) +
#     geom_edges(
#       aes(color = Color, linewidth = Value),
#       arrow = arrow(length = unit(5, "pt"), type = "closed")
#     ) +
#     geom_nodes(color = "gray", size = 3) +
#     scale_color_identity() +
#     theme_void() +
#     geom_nodelabel_repel(aes(label = name), size = 2) +
#     labs(title = "Beetle Co-occurrence Network")
#   } else {
#     cat("No significant beetle associations detected at threshold 0.1")
# }
#   
# ## Network visualisation Plants (note no if statement?)
# 
# g_plants <- graph_from_data_frame(assoc_net_plants_filtered, directed = TRUE)
# 
# E(g_plants)$Value <- assoc_net_plants_filtered$mean_strength
# E(g_plants)$Color <- ifelse(
#   assoc_net_plants_filtered$direction == "negative",
#   "blue",
#   "red"
# )
# 
# # Use circular layout for better visualization
# circle_layout <- layout_in_circle(g_plants)
# gg_plants <- ggnetwork(g_plants, layout = circle_layout)
# 
# ggplot(gg_plants, aes(x = x, y = y, xend = xend, yend = yend)) +
#   geom_edges(
#     aes(color = Color, linewidth = Value),
#     curvature = 0.1,
#     arrow = arrow(length = unit(5, "pt"), type = "closed")
#   ) +
#   geom_nodes(color = "gray", size = degree(g_plants, mode = "out") + 2) +
#   scale_color_identity() +
#   theme_void() +
#   geom_nodelabel_repel(
#     aes(label = name),
#     size = 2,
#     colour = "white",
#     fill = "grey36"
#   ) +
#   labs(
#     title = "Plant Co-occurrence Network",
#     subtitle = "Strong positive associations between epiphytic species"
#   )
# 
# 
# 
# ggplot(performance_summary, aes(x = Model, y = MCC, fill = Community)) +
#   geom_bar(stat = "identity", position = "dodge") +
#   theme_bw() +
#   labs(
#     title = "Model Performance Comparison Across Communities",
#     x = "Model Type",
#     y = "Matthews Correlation Coefficient"
#   ) +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1))
# 
timings$end <- Sys.time()
print(timings)
```
