---
title: "mrIML"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mrIML}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`mrIML` is a package for interpretable machine learning in R, specifically designed for multivariate (multi-response) data. It provides tools for building and interpreting multi-response models using the [`tidymodels`](https://www.tidymodels.org/start/) syntax; making it very useful for applications in community ecology and ecological genomics.

```{r setup}
library(mrIML)
library(tidymodels)
library(flashlight)
library(future)
library(ggplot2)
library(patchwork)

set.seed(7007)
```

Here we demonstrate a basic `mrIML` workflow using the `Bird.parasites` data set from the `MRFcov` package. This data set contains binary occurrences of four blood parasite species--*Hzosteropis, Hkillangoi, Plas, and Microfilaria*--in New Caledonian birds and a scaled continuous variable representing host relative abundance: *scale.prop.zos*.

```{r load-data}
data <- MRFcov::Bird.parasites
head(data)
```

To fit a multi-response models, we must first separate the data into the responses, the variables we wish to predict, and the covariates, the variables used only as predictors.

```{r prep-data}
# Responses
Y <- select(data, "Hzosteropis", "Hkillangoi", "Plas", "Microfilaria")
# Covariates
X <- select(data, "scale.prop.zos")
```

Many of the functions in the `mrIML` wrokflow that follows can be greatly sped up by using parallel processing. To do this, we just need to set up a cluster using the `future` package.

```{r setup-cluster, eval = FALSE}
plan(multisession, workers = 2)
```

# Fitting a mrIML model

The first step in the `mrIML` workflow is to define some models that we would like to compare. Thanks to `tidymodels`, we can define a wide range of models using a unified syntax. To demonstrate, we are going to set up two models to compare: a random forest model (RF) and a logistic regression (lm).

```{r define-tidymodels}
model_rf <- rand_forest(
  trees = 100, # 100 trees are set for brevity. Aim to start with 1000.
  mode = "classification",
  mtry = tune(),
  min_n = tune()
) %>%
  set_engine("randomForest")

model_lm <- logistic_reg() %>%
  set_engine("glm")
```

The models can then be fit as multi-response models using `mrIMLpredicts()`, which is the work horse of the `mrIML` package. `mrIMLpredicts()` fits the supplied tidymodel to each response variables. The `X1` argument can be used to supply any response variables that we would like to also use as predictors; which results in a co-ocurance model (in this case we use all the response variables).

```{r fit-mrIML, message = FALSE, echo=FALSE}
mrIML_rf <- mrIMLpredicts(
  X = X,
  Y = Y,
  X1 = Y,
  Model = model_rf,
  prop = 0.7,
  k = 5,
  racing = TRUE
)

mrIML_lm <- mrIMLpredicts(
  X = X,
  Y = Y,
  X1 = Y,
  Model = model_lm , 
  balance_data = 'no',
  prop = 0.6,
  k = 5,
  racing = FALSE
)
```

The resulting fitted `mrIML` model is a list object containing the original tidymodel, the data, and the fitted response models. The remainder of the functions in the `mrIML` package are aimed at interrogating, explaining, and comparing these fitted `mrIML` models.

# Model performance and comparison

`mrIMLperformance()` is another core function in `mrIML` used to extract performance metrics for each of the fitted response models inside the multi-response `mrIML` model.

```{r performance-mrIML, warning=FALSE}
perf_rf <- mrIML_rf %>%
  mrIMLperformance()
perf_rf$model_performance

perf_lm <- mrIML_lm %>%
  mrIMLperformance()
perf_rf$model_performance
```

These performance metrics can be compared between different `mrIML` model using the `mrPerformancePlot()` function.

```{r compare-mrIML}
perf_comp <- mrPerformancePlot(perf_rf, perf_lm)
perf_comp$performance_plot + perf_comp$performance_diff_plot
```

# Interogating models

To explore the effect of predictors on the different response models, `mrIML` works closely with the [`flashlight`](https://mayer79.github.io/flashlight/index.html) package, particularly their implementation of `flashlight::light_profile()` to generate partial dependency curves (PDs) for model agnostic comparisons. The `mrFlashlight()` function can be used to convert a `mrIML` model into a `flashlight` object in order to apply most of the `flashlight` functions.

```{r}
fl_rf <- mrIML_rf %>%
  mrFlashlight()

fl_rf$Microfilaria %>%
  light_profile(data = data, v = "scale.prop.zos") %>%
  plot() +
  ggtitle("Effect of scale.prop.zos on Microfilaria") +
  theme_bw()
```

`mrCovar()` is a wrapper around `mrFlashlight()` and `flashlight::light_profile()` to quickly compare the effect of a predictive covariate across the many response models.

```{r}
PD_scale.prop.zos_rf <- mrIML_rf %>%
  mrCovar(var = "scale.prop.zos", sdthresh = 0)

PD_scale.prop.zos_rf[[1]] /
  PD_scale.prop.zos_rf[[2]] /
  PD_scale.prop.zos_rf[[3]] +
  plot_layout(axis = "collect")
```

Since PDs are model agnostic we can use them for any kind of predictive model. For example we can compare the random forest model results above directly with the linear model bellow. We could even compare a Neural Net if we had enough data to fit it!

```{r}
PD_scale.prop.zos_lm <- mrIML_lm %>%
  mrCovar(var = "scale.prop.zos", sdthresh = 0)

PD_scale.prop.zos_lm[[1]] /
  PD_scale.prop.zos_lm[[2]] /
  PD_scale.prop.zos_lm[[3]] +
  plot_layout(axis = "collect")
```

The variability in the response probability across a PD curve can be used to rank variable importance in each response model. `mrVip()` compares variable importance across the response models inside a `mrIML` model.

```{r}
vip_rf <- mrIML_rf %>%
  mrVip()

vip_rf[[3]]
```

If there are many response models and/or predictive covariates, the `threshold`, `global_top_var`, and `local_top_var` arguments can be used to filter the variable importance results that are plotted by `mrVip`. The `taxa` argument can also be used to plot a desirable response model that may not fit the filtering criteria. See `?mrVip()` for the specifics of these arguments.

# Bootstrapping for uncertainty

To quantify uncertainty in a `mrIML` model, the function `mrBootstrap()` bootstraps the PDs. The output can then be passed to `mrPdPlotBootstrap()` to visualize the PDs.

```{r}
mrIML_boot_rf <- mrIML_rf %>%
  mrBootstrap()

mrPdPlotBootstrap(
  mrIML_rf,
  mrBootstrap_obj = mrIML_boot_rf,
  target = "Plas",
  global_top_var = 4
)[[2]]
```

After bootstrapping the PDs, the bootstrap information can be supplied to `mrVip()` to include uncertainty bounds in the plots.

```{r}
vip_boot_rf <- mrVip(
  mrIMLobj = mrIML_rf,
  mrBootstrap_obj = mrIML_boot_rf
)

vip_boot_rf[[3]]
```

# Checking for interactions

Another helpful investigative tool is `mrInteraction()`. This function quickly identifies which response models have strong evidence of interactions between predictors, which predictors have the most interactive effects and which interactions are the strongest.

```{r}
interactions_rf <- mrIML_rf %>%
  mrInteractions(feature = "Plas", num_bootstrap = 10)

interactions_rf[[1]] /
  interactions_rf[[2]] /
  interactions_rf[[3]]
```
