---
title: "Example 3: Forest Communities"
author: "Simon Riley, Ryan Leadbetter, Nick Fountain-Jones"
date: "2026-04-01"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This case study demonstrates *MrIML 2.0* on a larger data set with
multiple taxonomic groups across a forest successional gradient. We
analyze 144 plots established across forest age categories:

-   **Mature forest**: Unharvested with no fire disturbance for ≥60
    years
-   **\~45 years post-harvest**
-   **\~27 years post-harvest**
-   **\~7 years post-harvest**

At each plot, we recorded ground-active beetles, plants, and birds,
along with environmental covariates including coarse wood debris, leaf
litter depth, and soil variables (pH, conductivity).

The following is a series of random forests models to demonstrate different ways
models can be interpreted including use of co-occurrences, interactions and 
highlighting of variable importance.


The process within this vignette is:
-   Importing Data
-   Data Preparation
-   Model Creation
-   Bootstrapping Analysis
-   Variable Importance Assessment
-   Creation of co-occurrence networks

```{r libraryLoading, echo=TRUE, include=FALSE}
library(rsample, warn.conflicts=F, quietly=TRUE)
library(embed, warn.conflicts=F, quietly=TRUE)
library(dplyr, warn.conflicts=F, quietly=TRUE)
library(randomForest, warn.conflicts=F, quietly=TRUE)
library(tidyverse, warn.conflicts=F, quietly=TRUE)
library(tidymodels, warn.conflicts=F, quietly=TRUE)
library(mrIML, warn.conflicts=F, quietly=TRUE)
library(future, warn.conflicts=F, quietly=TRUE)
library(igraph)

library(patchwork, warn.conflicts=F, quietly=TRUE)
library(ggnetwork)
# Set up parallel processing

n_cores <- parallel::detectCores()
options(future.globals.maxSize = +Inf) # This is dirty, really needs optimisation
plan("multisession", workers = n_cores-2)

# Value for bootstrapping ()
bsNum <- 5
set.seed(1)
timings <- list()
```

## Data Loading

Data is supplied with this case study in a single rdata file. There are
4 data-frames, one containing all the environmental predictors and one for
each of the observations types (beetles, birds and plants).


```{r Data Loading, echo=TRUE}

currentDir <- getwd()
currentDir <- "C:/Users/Nebula PC/Documents/mrIML_HR"
sourceFile <- paste0(currentDir,"/R/mrIMLpredicts.R") 
source(sourceFile) #This needs to be removed when file is complete
loadFile <- paste0(currentDir,"/data/XYForrestData.rdata")
load(loadFile)

timings <- list()

```

## Data Preparation

Derived columns are removed and variables are arranged into X
(predictor), Y (response) and X1 (co-predictor) for each of beetles,
birds and plants.Data is converted to presence absence data.

Response variables are filtered using a mrIML function to remove species that
are either too common (over 70% prevalence) or too rare (less than 20% 
prevalence) to provide value in the analysis.


```{r Data Preparation, echo=TRUE}
timings$data$start <- Sys.time()

X_all <- data_site %>%
  dplyr::select(
    -any_of("Beetles_Mat_Ratio"),
    -any_of("Birds_Mat_Ratio"),
    -any_of("Slope_F"),
    -any_of("Aspect_Cor"),
    -any_of("Vegetation_cover"),
    -contains("Bryo"),
    -contains("Moss"),
    -contains('Plant')    
  )

X_beetles <- data_site %>%
  dplyr::select(
    -any_of("Beetles_Mat_Ratio"),
    -any_of("Slope_F"),
    -any_of("Aspect_Cor")    
  )

X_birds <- X_beetles

X_plants <- data_site %>%
  select(
    -contains("Bryo"),
    -contains("Moss"),
    -contains('Plant'),
    -Vegetation_cover)


Y_beetles <- data_beetles %>%
  dplyr::select(-Site, -Distance, -Transect, -Age) %>%
  dplyr::mutate_all(~ ifelse(. > 0, 1, .)) %>%
  mrIML::filterRareCommon(lower = 0.2, higher = 0.7)


Y_birds <- data_birds %>%
  dplyr::select(-Site, -Distance, -Transect, -Age) %>%
  mutate_all(~ ifelse(. > 0, 1, .)) %>%
  mrIML::filterRareCommon(lower = 0.2, higher = 0.7)

Y_plants <- data_plants %>%
  dplyr::select(-Site, -Distance, -Transect, -Age) %>%
  mutate_all(~ ifelse(. > 0, 1, .))%>%
  mrIML::filterRareCommon(lower = 0.2, higher = 0.7)

X1_beetles <- Y_beetles
X1_birds <- Y_birds
X1_plants <- Y_plants

Y_all <- cbind(Y_beetles,Y_birds,Y_plants)

timings$data$finish <- Sys.time()

```
## Data Summaries {.tabset}
### Data

Each of these tabs contain a summary either the raw or filtered version of the
data that will be used in the models. 

### Beetles Raw
```{r Beetles Raw Table}
dim(data_beetles)
summary(data_beetles)
```
### Beetles Filtered
```{r Beetles Filtered Table}
dim(Y_beetles)
summary(Y_beetles)
```
### Plants Raw
```{r Plants Raw Table}
dim(data_plants)
summary(data_plants)
```
### Plants Filtered
```{r Plants Filtered Table}
dim(Y_plants)
summary(Y_plants)
```
### Birds Raw
```{r Birds Raw Table}
dim(data_birds)
summary(data_birds)
```
### Birds Filtered
```{r Birds Filtered Table}
dim(Y_birds)
summary(Y_birds)
```

## Model Building {.tabset}

### Model Training

A common model tidy model specification is created.

```{r building models, echo = TRUE, results=TRUE}
timings$modelTemplate$start <- Sys.time()
# Model Spec with Tuning Parameters
rf_spec <- rand_forest(
  mtry = tune(),
  min_n = tune(),
  trees = 1000
) %>%
  set_mode("classification") %>%
  set_engine("randomForest")

```

Models are trained for environment without co-occurrence, co-occurrence
without the environmental variable, a combined model and a combined model
grouped by site.


### Beetles

The output here is 4 models for prediction of the presence of beetles.

```{r Training Beetle, echo= TRUE, results="hide", message= FALSE, warning=FALSE}
timings$beetleTrain$start <- Sys.time()

## Initial models for all response variables, environment, co-occurance and all data.
# Model 1: Environment only

yhats_beetles_env <- mrIMLpredicts(
  X = X_beetles,
  Y = Y_beetles,
  Model = rf_spec,
  balance_data = 'no',
  tune_grid_size = 5,
  prop = 0.6,
  k = 5,
  racing = TRUE
)


# Model 2: Co-occurrence only
yhats_beetles_cooccur <- mrIMLpredicts(
  Y = Y_beetles,
  X1 = X1_beetles,
  Model = rf_spec,
  balance_data = 'no',
  tune_grid_size = 5,
  prop = 0.6,
  k = 5,
  racing = TRUE
)

# Model 3: Combined
yhats_beetles_combined <- mrIMLpredicts(
  X = X_beetles,
  Y = Y_beetles,
  X1 = X1_beetles,
  Model = rf_spec,
  balance_data = 'no',
  prop = 0.6,
  k = 5,
  racing = FALSE
)

# Model 4: Combined
yhats_beetles_combined_H <- mrIMLpredicts(
  X = X_beetles,
  Y = Y_beetles,
  X1 = X1_beetles,
  Model = rf_spec,
  balance_data = 'no',
  prop = 0.6,
  k = 5,
  racing = FALSE,
  hierarchy = "Site"
)
timings$beetleTrain$finish <- Sys.time()

```

### Birds

The output here is 4 models for prediction of the presence of birds

```{r Training Birds, echo= TRUE, results="hide", message= FALSE, warning=FALSE}

timings$birdTrain$start <- Sys.time()

yhats_birds_env <- mrIMLpredicts(
  X = X_birds,
  Y = Y_birds,
  Model = rf_spec,
  balance_data = 'no',
  tune_grid_size = 5,
  prop = 0.6,
  k = 5,
  racing = TRUE
)


# Model 2: Co-occurrence only
yhats_birds_cooccur <- mrIMLpredicts(
  Y = Y_birds,
  X1 = X1_birds,
  Model = rf_spec,
  balance_data = 'no',
  tune_grid_size = 5,
  prop = 0.6,
  k = 5,
  racing = TRUE
)

# Model 3: Combined
yhats_birds_combined <- mrIMLpredicts(
  X = X_birds,
  Y = Y_birds,
  X1 = X1_birds,
  Model = rf_spec,
  balance_data = 'no',
  prop = 0.6,
  k = 5,
  racing = FALSE
)

# Model 4: Combined
yhats_birds_combined_H <- mrIMLpredicts(
  X = X_birds,
  Y = Y_birds,
  X1 = X1_birds,
  Model = rf_spec,
  balance_data = 'no',
  prop = 0.6,
  k = 5,
  racing = FALSE,
  hierarchy = "Site"
)
timings$birdTrain$finish <- Sys.time()

```

### Plants

The output here is 4 models for prediction of the presence of plants.

```{r Training Plants, echo= TRUE, results="hide", message= FALSE, warning=FALSE}

timings$plantTrain$start <- Sys.time()

yhats_plants_env <- mrIMLpredicts(
  X = X_plants,
  Y = Y_plants,
  Model = rf_spec,
  balance_data = 'no',
  tune_grid_size = 5,
  prop = 0.6,
  k = 5,
  racing = TRUE
)


# Model 2: Co-occurrence only
yhats_plants_cooccur <- mrIMLpredicts(
  Y = Y_plants,
  X1 = X1_plants,
  Model = rf_spec,
  balance_data = 'no',
  tune_grid_size = 5,
  prop = 0.6,
  k = 5,
  racing = TRUE
)

# Model 3: Combined
yhats_plants_combined <- mrIMLpredicts(
  X = X_plants,
  Y = Y_plants,
  X1 = X1_plants,
  Model = rf_spec,
  balance_data = 'no',
  prop = 0.6,
  k = 5,
  racing = FALSE
)

# Model 4: Combined
yhats_plants_combined_H <- mrIMLpredicts(
  X = X_plants,
  Y = Y_plants,
  X1 = X1_plants,
  Model = rf_spec,
  balance_data = 'no',
  prop = 0.6,
  k = 5,
  racing = FALSE,
  hierarchy = "Site"
)
timings$plantTrain$finish <- Sys.time()

```

## Bootstrapping

Bootstrapping is done to the combined data model and the hierarchical
data model to get a better estimate of the population properties.

```{r Bootstrap, echo=TRUE, results="hide", message= FALSE, warning=FALSE}
timings$bs_beetle$start = Sys.time()

bs_beetles <- mrBootstrap(yhats_beetles_combined, num_bootstrap = bsNum)
plan("sequential") 
bs_beetles_H <- mrBootstrap(yhats_beetles_combined_H, num_bootstrap = bsNum)
plan("multisession", workers = n_cores-2)
timings$bs_beetle$finish = Sys.time()

timings$bs_bird$start = Sys.time()
bs_birds <- mrBootstrap(yhats_birds_combined, num_bootstrap = bsNum)
plan("sequential")
bs_birds_H <- mrBootstrap(yhats_birds_combined_H, num_bootstrap = bsNum)
plan("multisession", workers = n_cores-2)
timings$bs_bird$finish = Sys.time()

timings$bs_plants$start = Sys.time()
bs_plants <- mrBootstrap(yhats_plants_combined, num_bootstrap = bsNum)

plan("sequential")
bs_plants_H <- mrBootstrap(yhats_plants_combined_H, num_bootstrap = bsNum)
plan("multisession", workers = n_cores-2)

timings$bs_plants$finish = Sys.time()
```

## Checking Variable Importance

Using the bootstrapping information variables are analysed for predictive power.

```{r Variable Importance Checks, echo = TRUE, warning=FALSE}
timings$vip_beetles$start <- Sys.time()
# Variable importance
vi_beetles <- mrVip(
  yhats_beetles_combined,
  mrBootstrap_obj = bs_beetles,
  threshold = 0.5,
  global_top_var = 10
)

# Variable importance
vi_beetles_H <- mrVip(
  yhats_beetles_combined_H,
  mrBootstrap_obj = bs_beetles_H,
  threshold = 0.5,
  global_top_var = 10
)
timings$vip_beetles$finish <- Sys.time()

timings$vip_plants$start <- Sys.time()
vi_plants <- mrVip(
  yhats_plants_combined,
  mrBootstrap_obj = bs_plants,
  threshold = 0.5,
  global_top_var = 10
)

vi_plants_H <- mrVip(
  yhats_plants_combined_H,
  mrBootstrap_obj = bs_plants_H,
  threshold = 0.5,
  global_top_var = 10
)
timings$vip_plants$finish <- Sys.time()

```

## Checking for interactions

Using the mrInteractions function the strength of interactions for a specific 
species. 

```{r Checking for interations, echo=TRUE}
timings$interactions$start <- Sys.time()

# Analyze interactions in beetle community
int_beetles <- mrInteractions(
  yhats_beetles_combined,
  num_bootstrap = bsNum,
  feature = 'Decilaus_nigronotatus',
  top_int = 10
)

int_beetles_H <- mrInteractions(
  yhats_beetles_combined_H,
  num_bootstrap = bsNum,
  feature = 'Decilaus_nigronotatus',
  top_int = 10
)

# Analyze interactions in beetle community
int_plants <- mrInteractions(
  yhats_plants_combined,
  num_bootstrap = bsNum,
  feature = 'Nothcunn',
  top_int = 10
)

int_plants_H <- mrInteractions(
  yhats_plants_combined_H,
  num_bootstrap = bsNum,
  feature = 'Nothcunn',
  top_int = 10
)

timings$interactions$end <- Sys.time()
```

## Evaluation of performance

Model outputs are evaluated.

```{r Evaluation of performance, echo=TRUE, warning = FALSE}
## Evaluate model performance beetles
ModelPerf_beetles_env <- mrIMLperformance(yhats_beetles_env)
ModelPerf_beetles_cooccur <- mrIMLperformance(yhats_beetles_cooccur)
ModelPerf_beetles_combined <- mrIMLperformance(yhats_beetles_combined)
ModelPerf_beetles_combined_H <- mrIMLperformance(yhats_beetles_combined_H)

perf_comparison_beetles_env <- mrPerformancePlot(
  ModelPerf1 = ModelPerf_beetles_env,
  ModelPerf2 = ModelPerf_beetles_combined_H,
  mode = "classification"
)

perf_comparison_beetles_cooccur <- mrPerformancePlot(
  ModelPerf1 = ModelPerf_beetles_cooccur,
  ModelPerf2 = ModelPerf_beetles_combined_H,
  mode = "classification"
)

perf_comparison_beetles_combined <- mrPerformancePlot(
  ModelPerf1 = ModelPerf_beetles_combined,
  ModelPerf2 = ModelPerf_beetles_combined_H,
  mode = "classification"
)

# Summary table
all_performance_beetles <- data.frame(
  Model = c("Environment only", "Co-occurrence only", "Combined","Hierarchy"),
  MCC = c(
    ModelPerf_beetles_env[[2]],
    ModelPerf_beetles_cooccur[[2]],
    ModelPerf_beetles_combined[[2]],
    ModelPerf_beetles_combined_H[[2]]
  )
)

## Evaluate model performance beetles
ModelPerf_plants_env <- mrIMLperformance(yhats_plants_env)
ModelPerf_plants_cooccur <- mrIMLperformance(yhats_plants_cooccur)
ModelPerf_plants_combined <- mrIMLperformance(yhats_plants_combined)
ModelPerf_plants_combined_H <- mrIMLperformance(yhats_plants_combined_H)


# Summary table
all_performance_plants <- data.frame(
  Model = c("Environment only", "Co-occurrence only", "Combined","Hierarchy"),
  MCC = c(
    ModelPerf_plants_env[[2]],
    ModelPerf_plants_cooccur[[2]],
    ModelPerf_plants_combined[[2]],
    ModelPerf_plants_combined_H[[2]]
  )
)

```

## Co-occurance Networks

```{r Co-occurance Networks, echo=TRUE}

timings$CN$start = Sys.time()

# Co-occurrence network
filterStrength <- 0.1

assoc_net_plants <- mrCoOccurNet(bs_plants)
assoc_net_plants_filtered <- assoc_net_plants %>%
  filter(mean_strength > filterStrength)


assoc_net_plants_H <- mrCoOccurNet(bs_plants_H)
assoc_net_plants_filtered_H <- assoc_net_plants_H %>%
  filter(mean_strength > filterStrength)

timings$CN$end = Sys.time()

```

## Importance Summary

From all this information we can get output tables and information on the most important predictors for beetles.

```{r Table Outputs, echo = TRUE}
all_performance_beetles
vi_beetles[[3]]
vi_beetles_H[[3]]
```


## Importance Summary

From all this information we can get output tables and information on the most important predictors for beetles.

```{r Table Outputs 2, echo = TRUE}
all_performance_plants
vi_plants[[3]]
vi_plants_H[[3]]
```

## Plots

From the package there are some plots that are prepared.

# Performance Comparison

Initially we look at the performance comparison plots

```{r Plots - performance comparision, echo=TRUE, warning= FALSE}

perf_comb=list(env = perf_comparison_beetles_env$performance_plot,
               # cooccr = perf_comparison_beetles_cooccur$performance_plot,
               comb = perf_comparison_beetles_combined$performance_plot)


patchwork::wrap_plots(perf_comb)

perf_comparison_beetles_combined$performance_plot +
  labs(title = "Community Model Performance")
```

# Importance Plots

Looking at importance plots between the models with and without
hierarchical grouping we can see how the importance of different
predictors is made more clear for this separation.

```{r Plots - Importance}


int_beetles[[1]] /
  int_beetles[[2]] /
  int_beetles[[3]] +
  plot_annotation(title = "Community Interactions")


int_beetles_H[[1]] /
  int_beetles_H[[2]] /
  int_beetles_H[[3]] +
  plot_annotation(title = "Community Interactions with Hierarchy")



int_plants[[1]] /
  int_plants[[2]] /
  int_plants[[3]] +
  plot_annotation(title = "Plant Community Interactions")


int_plants_H[[1]] /
  int_plants_H[[2]] /
  int_plants_H[[3]] +
  plot_annotation(title = "Plant Community Interactions")

```

# Network Visualisation

Using the associations calculated above they can be visualised as
follows:

```{r Plots - Network, echo = TRUE}
## Network visualization Beetles
g_plants <- graph_from_data_frame(assoc_net_plants_filtered, directed = TRUE)
g_plants_H <- graph_from_data_frame(assoc_net_plants_filtered_H, directed = TRUE)
## Network visualization Beetles
if (length(g_plants) > 0) {

  
  E(g_plants)$Value <- assoc_net_plants_filtered$mean_strength
  E(g_plants)$Color <- ifelse(
    assoc_net_plants_filtered$direction == "negative",
    "blue",
    "red"
  )
  
  # Use circular layout for better visualization
  circle_layout <- layout_in_circle(g_plants)
  gg_plants <- ggnetwork(g_plants, layout = circle_layout)
  
  ggplot(gg_plants, aes(x = x, y = y, xend = xend, yend = yend)) +
    geom_edges(
      aes(color = Color, linewidth = Value),
      curvature = 0.1,
      arrow = arrow(length = unit(5, "pt"), type = "closed")
    ) +
    geom_nodes(color = "gray", size = degree(g_plants, mode = "out") + 2) +
    scale_color_identity() +
    theme_void() +
    geom_nodelabel_repel(
      aes(label = name),
      size = 2,
      colour = "white",
      fill = "grey36"
    ) +
    labs(
      title = "Plant Co-occurrence Network",
      subtitle = "Strong positive associations between epiphytic species"
    )
  } else {
    cat(paste("No significant beetle associations detected at threshold ",filterStrength))
}

## Network visualization Hierarchy
if (length(g_plants_H) > 0) {
  
  E(g_plants_H)$Value <- assoc_net_plants_filtered_H$mean_strength
  E(g_plants_H)$Color <- ifelse(
    assoc_net_plants_filtered_H$direction == "negative",
    "blue",
    "red"
  )
  
  # Use circular layout for better visualization
  circle_layout_H <- layout_in_circle(g_plants_H)
  gg_plants_H <- ggnetwork(g_plants_H, layout = circle_layout_H)
  
  ggplot(gg_plants_H, aes(x = x, y = y, xend = xend, yend = yend)) +
    geom_edges(
      aes(color = Color, linewidth = Value),
      curvature = 0.1,
      arrow = arrow(length = unit(5, "pt"), type = "closed")
    ) +
    geom_nodes(color = "gray", size = degree(g_plants_H, mode = "out") + 2) +
    scale_color_identity() +
    theme_void() +
    geom_nodelabel_repel(
      aes(label = name),
      size = 2,
      colour = "white",
      fill = "grey36"
    ) +
    labs(
      title = "Plant Co-occurrence Network",
      subtitle = "Strong positive associations between epiphytic species"
    )
  } else {
    cat(paste("No significant beetle associations detected at threshold ",filterStrength))
}


```


```{r Timing, echo = TRUE}
timings

```

### Performance Patterns

Based on the expected results from the manuscript:

1.  **Environment-only models performed best** across all taxonomic
    groups
2.  **Plant communities** showed highest predictive performance (MCC ≈
    0.45)
3.  **Bird and beetle communities** showed similar, lower performance
    (MCC ≈ 0.21)
4.  **Co-occurrence patterns had minimal impact** on model performance


### Taxonomic Group Differences

#### Beetles

-   **Key predictors**: Altitude, soil characteristics, microclimate
    variables (LAI, heat load index)
-   **Limited associations**: Only one significant association detected
    (two *Decilaus* weevil species)
-   **Interpretation**: Species largely respond independently to
    environmental gradients

#### Birds

-   **Key predictors**: Altitude dominates
-   **No significant associations**: Species presences appear
    independent
-   **Low explanatory power**: Suggests unmeasured structural habitat
    variables are important
-   **Example**: Only Grey Fantail (*Rhipidura albiscapa*) showed modest
    improvement with co-occurrence data

#### Plants

-   **Key predictors**: Forest age (succession), LAI, Easting gradient\
-   **Strongest associations**: Multiple positive associations,
    particularly among epiphytic species
-   **Best performance**: Clear successional patterns from early-seral
    to shade-tolerant rainforest species
-   **Network structure**: Epiphytic ferns and bryophytes show strong
    co-occurrence (shared substrate requirements)

### Ecological Insights

Key plant associations detected:
- Large tree fern provides substrate for Rumohra adiantiformis
- Multiple epiphytic species co-occur on suitable substrates
- Succession drives community composition more than biotic interactions

Important environmental gradients:
- Altitude: Primary driver for beetles and birds
- Forest age: Critical for plant succession
- LAI and Easting: Important for beetles and plants
- Soil characteristics: Shape beetle communities


## Methodological Insights

### Model Selection Strategy

The comparison of three model types (environment-only,
co-occurrence-only, combined) reveals that:

1.  **Environmental filtering dominates** community assembly in these
    forest systems
2.  **Biotic interactions are secondary** but detectable in plant
    communities
3.  **Co-occurrence data alone** performs poorly, emphasizing the
    importance of environmental drivers

### Variable Selection Impact

Variable selection reduced dimensionality:
- Removed highly correlated predictors (>0.8 correlation)
- Eliminated predictors with minimal variance
- Retained ecologically meaningful gradients
- Improved model interpretability and performance


### Bootstrapping Benefits

The bootstrap analysis provides: - **Uncertainty estimation** for
variable importance - **Confidence intervals** for partial dependencies\
- **Robust network construction** from multiple model fits -
**Assessment of prediction stability**

### Hierarchical Considerations

The use of clustering by site enables the attributes of the site to 



## Comparison with Previous Studies

This analysis extends previous work (Fountain-Jones et al. 2015;
Hingston et al. 2014) by:

1.  **Unified framework**: Consistent methodology across taxonomic
    groups
2.  **Interaction quantification**: Explicit measurement of biotic vs.
    abiotic effects\
3.  **Non-linear relationships**: Capture of threshold and plateau
    effects
4.  **Uncertainty quantification**: Bootstrap confidence intervals
5.  **Network analysis**: Systematic co-occurrence network construction

### Novel Findings

-   **Asymmetric plant associations**: Direction of influence between
    epiphytic species
-   **Quantified interaction strength**: H-statistics reveal interaction
    importance
-   **Threshold identification**: Non-linear responses to environmental
    gradients
-   **Cross-community comparison**: Systematic differences in assembly
    mechanisms

## Management Implications

### Forest Succession

The strong performance of environment-only models suggests: - **Habitat
management** should focus on environmental conditions - **Successional
stage** is critical for plant community development - **Microclimate
variables** (LAI, heat load) are important for beetles - **Structural
complexity** may be key for bird communities (unmeasured)

### Conservation Priorities

Conservation implications:
1. Maintain environmental gradients across successional stages
2. Preserve old-growth characteristics (substrate for epiphytes)
3. Consider altitude and aspect in reserve design
4. Manage for microclimate heterogeneity
5. Monitor structural habitat development for bird communities


## Technical Considerations

### Model Performance Limitations

-   **Bird models**: Low explanatory power suggests missing predictors
-   **Rare species**: Filtered out due to insufficient data (prevalence
    thresholds)
-   **Temporal dynamics**: Cross-sectional data limits causal inference
-   **Scale effects**: Plot-level analysis may miss landscape-scale
    processes

### Future Improvements

1.  **Structural habitat measures**: LiDAR-derived metrics for bird
    habitat
2.  **Temporal replication**: Multiple sampling periods to assess
    stability
3.  **Functional traits**: Incorporation of species trait data
4.  **Phylogenetic relationships**: Account for evolutionary constraints
5.  **Spatial autocorrelation**: Explicit spatial modeling components

## Conclusions

This forest community analysis demonstrates that **environmental
filtering is the dominant force** shaping community assembly across
taxonomic groups, with **biotic interactions playing secondary roles**.
The *MrIML 2.0* framework successfully:

1.  **Quantifies relative importance** of environmental vs. biotic
    factors
2.  **Reveals taxonomic differences** in assembly mechanisms\
3.  **Identifies key environmental drivers** for each community
4.  **Constructs interpretable co-occurrence networks**
5.  **Provides uncertainty estimates** through bootstrapping

The results emphasize the importance of **maintaining environmental
heterogeneity** and **successional gradients** for forest biodiversity
conservation, while highlighting the need for **improved structural
habitat characterization** particularly for bird communities.




```{r Performance Patterns}


```

## Session Information

```{r session-info}
sessionInfo()
```



