---
title: "Hierarchy Absence Presence Example"
author: "Simon Riley, Ryan Leadbetter, Nick Fountain-Jones"
date: "2026-04-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
This case study demonstrates *MrIML 2.0* on a larger dataset with
multiple taxonomic groups across a forest successional gradient. We
analyze 144 plots established across forest age categories:

-   **Mature forest**: Unharvested with no fire disturbance for ≥60
    years
-   **\~45 years post-harvest**
-   **\~27 years post-harvest**
-   **\~7 years post-harvest**

At each plot, we recorded ground-active beetles, plants, and birds,
along with environmental covariates including coarse wood debris, leaf
litter depth, and soil variables (pH, conductivity).

```{r libraryLoading, echo=TRUE, include=FALSE}
library(rsample, warn.conflicts=F, quietly=TRUE)
library(embed, warn.conflicts=F, quietly=TRUE)
library(dplyr, warn.conflicts=F, quietly=TRUE)
library(randomForest, warn.conflicts=F, quietly=TRUE)
library(tidyverse, warn.conflicts=F, quietly=TRUE)
library(tidymodels, warn.conflicts=F, quietly=TRUE)
library(mrIML, warn.conflicts=F, quietly=TRUE)
library(future, warn.conflicts=F, quietly=TRUE)

library(patchwork, warn.conflicts=F, quietly=TRUE)

source("C:/Users/sriley0/OneDrive - University of Tasmania/RProjects/mrIML_HR/mrIML_HR/R/mrIMLpredicts.r") # This needs to be removed when updated

# Set up parallel processing
n_cores <- parallel::detectCores()
plan("multisession", workers = n_cores-2)

# Value for bootstrapping ()
bsNum <- 20
set.seed(1)
timings <- list()
```

## Data Loading

Data is supplied with this case study in a single rdata file. There are 4 
dataframes, one containing all the environmental elements and one for each of 
the observations types (beetles, birds and plants).

Derived columns are removed and variables are arranged into X (predictor),
Y (response) and X1 (co-predictor) for each of beetles, birds and plants. 



```{r data, echo=TRUE}
sourceFile <- paste0(currentDir,"/R/mrIMLpredicts.R")
source(sourceFile)
loadFile <- paste0(currentDir,"/data/XYForrestData.rdata")
load(loadFile)

timings <- list()

timings$data$start

X_all <- data_site %>%
  dplyr::select(
    -any_of("Beetles_Mat_Ratio"),
    -any_of("Birds_Mat_Ratio"),
    -any_of("Slope_F"),
    -any_of("Aspect_Cor"),
    -any_of("Vegetation_cover"),
    -contains("Bryo"),
    -contains("Moss"),
    -contains('Plant')    
  )

Y_beetles <- data_beetles %>%
  dplyr::select(-Site, -Distance, -Transect, -Age) %>%
  dplyr::mutate_all(~ ifelse(. > 0, 1, .)) %>%
  mrIML::filterRareCommon(lower = 0.2, higher = 0.7)


Y_birds <- data_birds %>%
  dplyr::select(-Site, -Distance, -Transect, -Age) %>%
  mutate_all(~ ifelse(. > 0, 1, .)) %>%
  mrIML::filterRareCommon(lower = 0.2, higher = 0.7)

Y_plants <- data_plants %>%
  dplyr::select(-Site, -Distance, -Transect, -Age) %>%
  mutate_all(~ ifelse(. > 0, 1, .))%>%
  mrIML::filterRareCommon(lower = 0.2, higher = 0.7)

X1_beetles <- Y_beetles
X1_birds <- Y_birds
X1_plants <- Y_plants

Y_all <- cbind(Y_beetles,Y_birds,Y_plants)

timings$data$finish

```

## Model Training

A common model tidy model specification is created.
 
```{r building models, echo = TRUE, results=TRUE}
timings$modelTrain$start
# Model Spec with Tuning Parameters
rf_spec <- rand_forest(
  mtry = tune(),
  min_n = tune(),
  trees = 1000
) %>%
  set_mode("classification") %>%
  set_engine("randomForest")

```

Models are trained for environment without co-occurance, co-occurance without the environmental variable, a combined model and a combined grouped by site.


```{r training models, echo= TRUE, results=FALSE}
timings$modelTrain_start

## Initial models for all response variables, environment, co-occurance and all data.
# Model 1: Environment only
yhats_all_env <- mrIMLpredicts(
  X = X_all,
  Y = Y_all,
  Model = rf_spec,
  balance_data = 'no',
  tune_grid_size = 5,
  prop = 0.6,
  k = 5,
  racing = TRUE
)

# Model 2: Co-occurrence only
yhats_all_cooccur <- mrIMLpredicts(
  Y = Y_all,
  X1 = X1_all,
  Model = rf_spec,
  balance_data = 'no',
  tune_grid_size = 5,
  prop = 0.6,
  k = 5,
  racing = TRUE
)

# Model 3: Combined
yhats_all_combined <- mrIMLpredicts(
  X = X_all,
  Y = Y_all,
  X1 = X1_all,
  Model = rf_spec, 
  balance_data = 'no',
  prop = 0.6,
  k = 5,
  racing = FALSE
)

# Model 4: Combined
yhats_all_combined_H <- mrIMLpredicts(
  X = X_all,
  Y = Y_all,
  X1 = X1_all,
  Model = rf_spec, 
  balance_data = 'no',
  prop = 0.6,
  k = 5,
  racing = FALSE,
  hierarchy = "Site"
)
timings$modelTrain$finish <- Sys.time()
```

## Bootstrapping 

Bootstrapping is done to the combined data model and the hierarchical data model
to get a better estimate of the population properties.

```{r Bootstrap, echo=TRUE}
timings$bs$start = Sys.time()

bs_beetles <- mrBootstrap(yhats_beetles_combined, num_bootstrap = bsNum)


plan("sequential")
bs_beetles_H <- mrBootstrap(yhats_beetles_combined_H, num_bootstrap = bsNum)
plan("multisession", workers = n_cores-2)

timings$bs$finish = Sys.time()

```


## Checking Variable Importance

```{r Variable Importance Checks, echo = TRUE}
timings$vip$start <- Sys.time()
# Variable importance
vi_all <- mrVip(
  yhats_all_combined,
  mrBootstrap_obj = bs_all,
  threshold = 0.5,
  global_top_var = 10
)


# Variable importance
vi_all_H <- mrVip(
  yhats_all_combined_H,
  mrBootstrap_obj = bs_all_H,
  threshold = 0.5,
  global_top_var = 10
)
timings$vip$end <- Sys.time()


```



## Evaluation of performance
Model outputs are evaluated. 

```{r Evaluation of performance, echo=TRUE}
## Evaluate model performance beetles
ModelPerf_all_env <- mrIMLperformance(yhats_all_env)
ModelPerf_all_cooccur <- mrIMLperformance(yhats_all_cooccur)
ModelPerf_all_combined <- mrIMLperformance(yhats_all_combined)
ModelPerf_all_combined_H <- mrIMLperformance(yhats_all_combined_H)

perf_comparison_all_env <- mrPerformancePlot(
  ModelPerf1 = ModelPerf_all_env,
  ModelPerf2 = ModelPerf_all_combined_H,
  mode = "classification"
)

perf_comparison_all_cooccur <- mrPerformancePlot(
  ModelPerf1 = ModelPerf_all_cooccur,
  ModelPerf2 = ModelPerf_all_combined_H,
  mode = "classification"
)

perf_comparison_all_combined <- mrPerformancePlot(
  ModelPerf1 = ModelPerf_all_combined,
  ModelPerf2 = ModelPerf_all_combined_H,
  mode = "classification"
)

# Summary table
all_performance <- data.frame(
  Model = c("Environment only", "Co-occurrence only", "Combined","Hierarchy"),
  MCC = c(
    ModelPerf_all_env[[2]],
    ModelPerf_all_cooccur[[2]],
    ModelPerf_all_combined[[2]],
    ModelPerf_all_combined_H[[2]]
  )
)





```

## Co-occurance Networks

```{r Co-occurance Networks, echo=TRUE}

timings$CN$start = Sys.time()

# Co-occurrence network
filterStrength <- 0.10

assoc_net_all <- mrCoOccurNet(bs_all)
assoc_net_all_filtered <- assoc_net_all %>%
  filter(mean_strength > filterStrength)


assoc_net_all_H <- mrCoOccurNet(bs_all_H)
assoc_net_all_filtered_H <- assoc_net_all_H %>%
  filter(mean_strength > filterStrength)



timings$CN$end = Sys.time()

```


## Summary Tables

From all this information we can get output tables

```{r Table Outputs, echo = TRUE}
cat("Performance:\n")
all_performance
vi_all[[3]]
```


## Plots
From the package there are some plots that are prepared.

# Performance Comparison 
Initially we look at the performance comparison plots

```{r Plots - performance comparision, echo=TRUE}

perf_comb=list(env = perf_comparison_all_env$performance_plot,
               cooccr = perf_comparison_all_cooccur$performance_plot,
               comb = perf_comparison_all_combined$performance_plot)


patchwork::wrap_plots(perf_comb)

perf_comparison_all_combined$performance_plot +
  labs(title = "Community Model Performance")
```

# Importance Plots

Looking at importance plots between the models with and without hierarchical 
grouping we can see how the importance of different predictors is made more clear
for this separation.

```{r Plots - Importance}
# Importance plots
vi_all[[3]]

# With hierarchy context
vi_all_H[[3]]


int_all[[1]] /
  int_all[[2]] /
  int_all[[3]] +
  plot_annotation(title = "Community Interactions")


int_all_H[[1]] /
  int_all_H[[2]] /
  int_all_H[[3]] +
  plot_annotation(title = "Community Interactions with Hierarchy")

```

# Network Visualisation

Using the associations calculated above they can be visualised as follows:

```{r Plots - Network, echo = TRUE}
## Network visualization Beetles

## Network visualization Beetles
if (nrow(assoc_net_all_filtered) > 0) {
  g_all <- graph_from_data_frame(
    assoc_net_all_filtered,
    directed = TRUE
  )

  E(g_all)$Value <- assoc_net_all_filtered$mean_strength
  E(g_all)$Color <- ifelse(
    assoc_net_all_filtered$direction == "negative",
    no = "blue",
    yes = "red"
  )

  gg_all <- ggnetwork(g_all)

  ggplot(gg_all, aes(x = x, y = y, xend = xend, yend = yend)) +
    geom_edges(
      aes(color = Color, linewidth = Value),
      arrow = arrow(length = unit(5, "pt"), type = "closed")
    ) +
    geom_nodes(color = "gray", size = 3) +
    scale_color_identity() +
    theme_void() +
    geom_nodelabel_repel(aes(label = name), size = 2) +
    labs(title = "Co-occurrence Network")
  } else {
    cat(paste("No significant beetle associations detected at threshold ",filterStrength))
}

## Network visualization Hierarchy
if (nrow(assoc_net_all_filtered_H) > 0) {
  g_all_H <- graph_from_data_frame(
    assoc_net_all_filtered_H,
    directed = TRUE
  )
  
  E(g_all_H)$Value <- assoc_net_all_filtered_H$mean_strength
  E(g_all_H)$Color <- ifelse(
    assoc_net_all_filtered_H$direction == "negative",
    no = "blue",
    yes = "red"
  )
  
  gg_all_H <- ggnetwork(g_all_H)
  
  ggplot(gg_all_H, aes(x = x, y = y, xend = xend, yend = yend)) +
    geom_edges(
      aes(color = Color, linewidth = Value),
      arrow = arrow(length = unit(5, "pt"), type = "closed")
    ) +
    geom_nodes(color = "gray", size = 3) +
    scale_color_identity() +
    theme_void() +
    geom_nodelabel_repel(aes(label = name), size = 2) +
    labs(title = "Co-occurrence Network")
} else {
  cat(paste("No significant beetle associations detected at threshold ",filterStrength))
}


```



# Conclusion

From here we can see that stuff happened.


```{r Timing, echo = TRUE}
timings



```
















