---
title: "Benchmarking and simulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benchmarking and simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  eval    = FALSE,
  warning = FALSE,
  message = FALSE
)
```

## Overview

This article benchmarks mrIML's co-occurrence network inference against three widely-used alternatives:

- **Hmsc** (Hierarchical Models of Species Communities) — a joint species distribution model that recovers residual associations via a latent-factor random effect, here using 95% posterior credible intervals to threshold edges.
- **SPIEC-EASI** — a sparse graphical model designed for compositional microbiome data.
- **SparCC** — a correlation method corrected for the compositional nature of count data.

All four methods are applied to presence–absence data generated from a generalised Lotka–Volterra (GLV) simulator across four ecological interaction scenarios: random, mutualistic, competitive, and mixed. Network recovery is evaluated with the Spearman Mantel test against the known (simulated) interaction matrix.

The simulation framework is adapted from the [mrIML paper supplementary material](https://rleadbett.github.io/mrIML-paper-supmat/sim-study/dummy.html). The key modification here is that the Hmsc network uses posterior-mean OmegaCor values, retaining only edges whose 95% credible intervals exclude zero.

---

## 1. Setup

```{r libraries}
# Data manipulation
library(here)
library(readxl)
library(tidyverse)
library(DT)

# Modelling
library(mrIML)
library(tidymodels)
library(future)
library(finetune)
library(flashlight)
library(Hmsc)      # Hierarchical Models of Species Communities
library(coda)      # MCMC diagnostics (Gelman-Rubin R-hat)

# Network construction and visualisation
library(igraph)
library(ggnetwork)
library(cowplot)
library(patchwork)
library(seqtime)   # GLV simulator (generateDataSet, glv)
library(vegan)     # Mantel test
library(SpiecEasi) # SPIEC-EASI and SparCC
```

```{r parallelism}
set.seed(109)

# Use all but two cores for mrIML's internal parallelism.
# The outer replicate loop remains sequential to avoid resource contention
# (mrIML already saturates available cores across species).
n_cores <- parallel::detectCores()
plan("multisession", workers = n_cores - 2)
```

---

## 2. Helper functions

### 2.1 Simulate presence–absence data

```{r fn-pa-data}
# Simulate a community from a GLV interaction matrix M.
# Returns both raw count data (needed by SPIEC-EASI / SparCC) and
# a presence-absence (PA) data frame (needed by mrIML / Hmsc).
make_presence_absence_data <- function(M, n_samples = 900, count = NULL, mode = 4) {
  if (is.null(count)) {
    # Default total count scales with network size
    count <- nrow(M) * 10000
  }

  # seqtime::generateDataSet returns a species × samples matrix
  raw_species_x_samples <- generateDataSet(n_samples, M, count = count, mode = mode)

  # Transpose to samples × species and binarise to presence (1) / absence (0)
  pa <- raw_species_x_samples |>
    t() |>
    as.data.frame() |>
    mutate(across(everything(), ~ ifelse(.x < 1, yes = 0, no = 1)))

  list(
    raw = raw_species_x_samples,  # species × samples; input format for SPIEC-EASI / SparCC
    pa  = pa                       # samples × species PA; input format for mrIML / Hmsc
  )
}
```

### 2.2 Fit an mrIML co-occurrence network

```{r fn-mrIML}
# Fit an mrIML random-forest co-occurrence network and convert it to an igraph object.
# Each species in Y is modelled as a function of all other species (X1 = Y).
# Edges are retained when mean bootstrapped association strength > mean_strength_cutoff.
fit_mrIML_network <- function(Y, model_rf, mean_strength_cutoff = 0.1) {
  X1 <- Y  # use all species as predictors for every other species

  yhats_rf_sim <- mrIMLpredicts(
    Y     = Y,
    X     = NULL,  # no environmental predictors in this simulation
    X1    = X1,
    Model = model_rf,
    prop  = 0.7,   # 70 / 30 train–test split per species model
    k     = 5,     # 5-fold cross-validation
    racing = FALSE
  )

  # Bootstrap across cross-validation folds to estimate edge uncertainty
  bs_sim   <- mrBootstrap(yhats_rf_sim)
  assoc_net <- mrCoOccurNet(bs_sim)

  # Convert association table to a weighted adjacency matrix.
  # Sign is encoded in mean_strength_dir (negative = competition / exclusion).
  mrIML_mat <- assoc_net |>
    filter(mean_strength > mean_strength_cutoff) |>
    mutate(
      mean_strength_dir = ifelse(
        direction == "negative",
        yes = -mean_strength,
        no  =  mean_strength
      ),
      # Strip the "sp" prefix added by mrCoOccurNet so row/col names are integers
      across(contains("taxa"), ~ sub("^sp", "", .x))
    ) |>
    graph_from_data_frame(
      directed = FALSE,
      vertices = sub("^sp", "", names(Y))
    ) |>
    as_adjacency_matrix(attr = "mean_strength_dir", sparse = FALSE)

  g_mrIML <- graph_from_adjacency_matrix(
    mrIML_mat,
    mode     = "undirected",
    weighted = TRUE,
    diag     = FALSE
  )

  list(
    yhats     = yhats_rf_sim,
    bootstrap = bs_sim,
    assoc_net = assoc_net,
    mat       = mrIML_mat,
    graph     = g_mrIML,
    edge_col  = ifelse(E(g_mrIML)$weight < 0, yes = "red", no = "blue"),
    edge_width = rep(0.4, ecount(g_mrIML))
  )
}
```

### 2.3 Fit an Hmsc model

```{r fn-hmsc-fit}
# Fit a probit Hmsc model with a sample-level random effect.
# The random effect captures residual species associations (OmegaCor) after
# accounting for the intercept-only fixed effects.
# MCMC settings (thin, samples, transient, nChains) trade accuracy for speed;
# use the defaults for final results and reduce for quick smoke tests.
fit_hmsc_model <- function(Y,
                           thin      = 100,
                           samples   = 5000,
                           transient = 20000,
                           nChains   = 4) {

  Y_hmsc <- as.data.frame(Y)

  # Remove any incidental row-index columns that would confuse Hmsc
  id_cols <- intersect(names(Y_hmsc), c("sample", "site", "row", "id", "X"))
  if (length(id_cols) > 0) {
    Y_hmsc <- dplyr::select(Y_hmsc, -dplyr::all_of(id_cols))
  }

  # Hmsc probit expects a numeric 0/1 matrix (rows = samples, columns = species)
  Y_hmsc <- as.matrix(dplyr::mutate(Y_hmsc, dplyr::across(dplyr::everything(), as.numeric)))

  # Drop incomplete rows to avoid row-number mismatches with XData / studyDesign
  keep_rows <- stats::complete.cases(Y_hmsc)
  Y_hmsc    <- Y_hmsc[keep_rows, , drop = FALSE]

  sample_ids <- paste0("sample_", seq_len(nrow(Y_hmsc)))
  rownames(Y_hmsc) <- sample_ids

  # Intercept-only fixed effects (no environmental covariates in this simulation)
  XData      <- data.frame(intercept = rep(1, nrow(Y_hmsc)), row.names = sample_ids)
  # One random level per sampling unit — this is what produces OmegaCor
  studyDesign <- data.frame(sample = factor(sample_ids), row.names = sample_ids)

  stopifnot(nrow(Y_hmsc) == nrow(XData))
  stopifnot(nrow(Y_hmsc) == nrow(studyDesign))

  rL <- HmscRandomLevel(units = levels(studyDesign$sample))

  m_hmsc <- Hmsc(
    Y           = Y_hmsc,
    XFormula    = ~1,
    XData       = XData,
    distr       = "probit",   # appropriate for binary presence-absence
    studyDesign = studyDesign,
    ranLevels   = list("sample" = rL)
  )

  sampleMcmc(
    m_hmsc,
    thin      = thin,
    samples   = samples,
    transient = transient,
    nChains   = nChains,
    verbose   = 0
  )
}
```

### 2.4 Check Hmsc MCMC convergence

```{r fn-hmsc-convergence}
# Compute the Gelman-Rubin R-hat for OmegaCor parameters.
# Values > 1.1 suggest the chains have not converged and results should be
# interpreted cautiously (or MCMC should be run longer).
check_hmsc_omega_convergence <- function(m_hmsc, rhat_cutoff = 1.1) {
  mpost <- convertToCodaObject(m_hmsc)
  psrf  <- gelman.diag(mpost$Omega[[1]], multivariate = FALSE)$psrf

  if (any(psrf[, "Point est."] > rhat_cutoff, na.rm = TRUE)) {
    warning(
      "Hmsc convergence concern: one or more Omega parameters have Rhat > ",
      rhat_cutoff,
      ". Interpret Hmsc network results cautiously."
    )
  }

  psrf
}
```

### 2.5 Build an Hmsc credible-interval network

```{r fn-hmsc-network}
# Construct an Hmsc residual-association network using 95% posterior credible intervals.
# Only edges whose 95% CrI for OmegaCor excludes zero are retained —
# this is more conservative than thresholding by the posterior mean alone.
make_hmsc_ci_network <- function(m_hmsc,
                                 mrIML_mat,
                                 ci           = c(0.025, 0.975),
                                 random_level = 1) {

  # getPostEstimate returns posterior mean and requested quantiles for OmegaCor
  post_omega <- getPostEstimate(
    m_hmsc,
    parName = "OmegaCor",
    r       = random_level,
    q       = ci
  )

  Omega_mean <- post_omega$mean
  # Hmsc does not set dimnames for OmegaCor; add species names manually
  dimnames(Omega_mean) <- list(m_hmsc$spNames, m_hmsc$spNames)

  # post_omega$q is a 3D array: quantile × species × species
  Omega_lwr <- post_omega$q[1, , ]
  Omega_upr <- post_omega$q[2, , ]
  dimnames(Omega_lwr) <- list(m_hmsc$spNames, m_hmsc$spNames)
  dimnames(Omega_upr) <- list(m_hmsc$spNames, m_hmsc$spNames)

  # Zero out any edge whose 95% CrI spans zero (i.e., not credibly non-zero)
  Omega_CI <- Omega_mean
  Omega_CI[Omega_lwr <= 0 & Omega_upr >= 0] <- 0
  diag(Omega_CI) <- 0

  # Match species names between the Hmsc matrix (uses "sp" prefix) and the
  # mrIML matrix (prefix stripped). Guard against naming convention mismatches.
  hmsc_all_names     <- colnames(Omega_mean)
  candidate_names    <- paste0("sp", rownames(mrIML_mat))
  hmsc_species_names <- intersect(candidate_names, hmsc_all_names)

  if (length(hmsc_species_names) == 0) {
    stop(
      "make_hmsc_ci_network: no species names in common between mrIML matrix (",
      paste(head(candidate_names), collapse = ", "),
      ") and Hmsc OmegaCor matrix (",
      paste(head(hmsc_all_names), collapse = ", "),
      "). Check species naming conventions."
    )
  }

  hmsc_mat <- Omega_CI[hmsc_species_names, hmsc_species_names, drop = FALSE]
  hmsc_lwr <- Omega_lwr[hmsc_species_names, hmsc_species_names, drop = FALSE]
  hmsc_upr <- Omega_upr[hmsc_species_names, hmsc_species_names, drop = FALSE]

  # Align row/col names with the mrIML matrix for downstream comparison
  rownames(hmsc_mat) <- colnames(hmsc_mat) <- rownames(mrIML_mat)
  rownames(hmsc_lwr) <- colnames(hmsc_lwr) <- rownames(mrIML_mat)
  rownames(hmsc_upr) <- colnames(hmsc_upr) <- rownames(mrIML_mat)

  g_hmsc <- graph_from_adjacency_matrix(
    hmsc_mat,
    mode     = "undirected",
    weighted = TRUE,
    diag     = FALSE
  )

  # Store the credible interval bounds and edge sign on the graph object
  if (ecount(g_hmsc) > 0) {
    edge_list       <- as_edgelist(g_hmsc)
    E(g_hmsc)$ci_lwr <- mapply(function(a, b) hmsc_lwr[a, b], edge_list[, 1], edge_list[, 2])
    E(g_hmsc)$ci_upr <- mapply(function(a, b) hmsc_upr[a, b], edge_list[, 1], edge_list[, 2])
    E(g_hmsc)$sign   <- ifelse(E(g_hmsc)$weight > 0, "positive", "negative")
  }

  list(
    post      = post_omega,
    mat       = hmsc_mat,
    lwr       = hmsc_lwr,
    upr       = hmsc_upr,
    graph     = g_hmsc,
    edge_col  = ifelse(E(g_hmsc)$weight < 0, yes = "red", no = "blue"),
    edge_width = rep(0.4, ecount(g_hmsc))
  )
}
```

### 2.6 Fit a SPIEC-EASI network

```{r fn-spiec}
# Fit a SPIEC-EASI sparse inverse-covariance network (MB = Meinshausen-Bühlmann).
# SPIEC-EASI was designed for compositional microbiome count data but performs
# well on other sparse ecological count matrices.
# raw_species_x_samples must be a species × samples matrix (seqtime convention).
fit_spiec_network <- function(raw_species_x_samples, Y) {
  sp_names       <- colnames(Y)                                    # e.g. "sp1", "sp3", ...
  raw_subset     <- raw_species_x_samples[sp_names, , drop = FALSE]
  sp_names_clean <- sub("^sp", "", sp_names)

  # SPIEC-EASI expects samples × features; transpose from species × samples
  se_fit   <- spiec.easi(t(raw_subset), method = "mb",
                         pulsar.params = list(rep.num = 20))
  beta_mat <- as.matrix(symBeta(getOptBeta(se_fit), mode = "lower"))
  diag(beta_mat) <- 0
  rownames(beta_mat) <- colnames(beta_mat) <- sp_names_clean

  g_spiec <- graph_from_adjacency_matrix(
    beta_mat,
    mode     = "undirected",
    weighted = TRUE,
    diag     = FALSE
  )

  list(
    mat       = beta_mat,
    graph     = g_spiec,
    edge_col  = ifelse(E(g_spiec)$weight < 0, "red", "blue"),
    edge_width = rep(0.4, ecount(g_spiec))
  )
}
```

### 2.7 Fit a SparCC network

```{r fn-sparcc}
# Fit a SparCC correlation network with bootstrap-based p-values.
# Edges with p > alpha are set to zero. SparCC corrects for the
# spurious correlations induced by compositionality in count data.
fit_sparcc_network <- function(raw_species_x_samples, Y,
                               alpha = 0.05, n_boot = 50) {
  sp_names       <- colnames(Y)
  raw_subset     <- raw_species_x_samples[sp_names, , drop = FALSE]
  sp_names_clean <- sub("^sp", "", sp_names)

  counts_mat  <- t(raw_subset)       # samples × features for SpiecEasi functions
  sparcc_out  <- SpiecEasi::sparcc(counts_mat)
  sparcc_boot <- SpiecEasi::sparccboot(counts_mat, R = n_boot)
  sparcc_pval <- SpiecEasi::pval.sparccboot(sparcc_boot)

  sparcc_cor <- sparcc_out$Cor
  sparcc_cor[sparcc_pval$pvals > alpha] <- 0   # retain only significant correlations
  diag(sparcc_cor) <- 0
  rownames(sparcc_cor) <- colnames(sparcc_cor) <- sp_names_clean

  g_sparcc <- graph_from_adjacency_matrix(
    sparcc_cor,
    mode     = "undirected",
    weighted = TRUE,
    diag     = FALSE
  )

  list(
    mat       = sparcc_cor,
    graph     = g_sparcc,
    edge_col  = ifelse(E(g_sparcc)$weight < 0, "red", "blue"),
    edge_width = rep(0.4, ecount(g_sparcc))
  )
}
```

### 2.8 Extract the simulated truth network

```{r fn-truth}
# Subset the GLV interaction matrix M to the species retained after
# mrIML's rare/common filtering. This gives the 'ground truth' sub-network
# against which all four methods are evaluated.
make_truth_network <- function(M, mrIML_mat) {
  matching_indices <- as.numeric(rownames(mrIML_mat))
  true_mat <- M[matching_indices, matching_indices, drop = FALSE]
  rownames(true_mat) <- colnames(true_mat) <- rownames(mrIML_mat)

  g_true <- graph_from_adjacency_matrix(
    true_mat,
    mode     = "undirected",
    weighted = TRUE,
    diag     = FALSE
  )

  list(
    mat       = true_mat,
    graph     = g_true,
    edge_col  = ifelse(E(g_true)$weight < 0, yes = "red", no = "blue"),
    edge_width = abs(E(g_true)$weight) * 5,  # edge width proportional to interaction strength
    layout    = layout_nicely(g_true)          # shared layout for all five panel plots
  )
}
```

### 2.9 Side-by-side network plot

```{r fn-plot}
# Plot all five networks (mrIML, Hmsc, SPIEC-EASI, SparCC, truth) side by side,
# using the simulated-truth layout so node positions are identical across panels.
# Red edges = negative (competitive / inhibitory) interactions;
# blue edges = positive (mutualistic / facilitative) interactions.
plot_five_networks <- function(mrIML_res, hmsc_res, spiec_res, sparcc_res, truth_res,
                               title_suffix = "") {
  par_old <- par(mfrow = c(1, 5), mar = c(1, 1, 2, 1))
  on.exit(par(par_old), add = TRUE)

  plot(mrIML_res$graph,  layout = truth_res$layout,
       edge.color = mrIML_res$edge_col,  edge.width = mrIML_res$edge_width,
       main = paste("mrIML", title_suffix))

  plot(hmsc_res$graph,   layout = truth_res$layout,
       edge.color = hmsc_res$edge_col,   edge.width = hmsc_res$edge_width,
       main = paste("Hmsc 95% CrI", title_suffix))

  plot(spiec_res$graph,  layout = truth_res$layout,
       edge.color = spiec_res$edge_col,  edge.width = spiec_res$edge_width,
       main = paste("SPIEC-EASI", title_suffix))

  plot(sparcc_res$graph, layout = truth_res$layout,
       edge.color = sparcc_res$edge_col, edge.width = sparcc_res$edge_width,
       main = paste("SparCC", title_suffix))

  plot(truth_res$graph,  layout = truth_res$layout,
       edge.color = truth_res$edge_col,  edge.width = truth_res$edge_width,
       main = paste("Simulated truth", title_suffix))
}
```

### 2.10 Compare inferred networks against the truth (Mantel test)

```{r fn-mantel}
# Use the Spearman Mantel test to measure the rank correlation between each
# method's inferred association matrix and the simulated interaction matrix.
# Matrices are on different scales, so Spearman (rank-based) correlation is used
# rather than comparing absolute values directly.
compare_with_truth <- function(true_mat, mrIML_mat, hmsc_mat, spiec_mat, sparcc_mat,
                               permutations = 999) {
  dist_true   <- as.dist(1 - true_mat)
  dist_mrIML  <- as.dist(1 - mrIML_mat)
  dist_hmsc   <- as.dist(1 - hmsc_mat)
  dist_spiec  <- as.dist(1 - spiec_mat)
  dist_sparcc <- as.dist(1 - sparcc_mat)

  list(
    mrIML      = vegan::mantel(dist_true, dist_mrIML,  method = "spearman", permutations = permutations),
    Hmsc       = vegan::mantel(dist_true, dist_hmsc,   method = "spearman", permutations = permutations),
    SPIEC_EASI = vegan::mantel(dist_true, dist_spiec,  method = "spearman", permutations = permutations),
    SparCC     = vegan::mantel(dist_true, dist_sparcc, method = "spearman", permutations = permutations)
  )
}
```

### 2.11 Orchestrate one complete simulation scenario

```{r fn-run-scenario}
# Run a single simulation scenario end-to-end:
#   1. Generate a GLV interaction network of the specified type
#   2. Simulate presence-absence community data from that network
#   3. Filter rare / ubiquitous species
#   4. Fit all four co-occurrence models
#   5. Plot inferred vs. truth networks
#   6. Evaluate recovery with Mantel tests
run_scenario <- function(scenario,
                         interaction_type,
                         interact_str_max,
                         network_size        = 20,
                         k_average           = 4,
                         mix_compt_ratio     = 0.5,
                         self_regulation     = NULL,
                         model_rf,
                         hmsc_thin           = 50,
                         hmsc_samples        = 1000,
                         hmsc_transient      = 5000,
                         hmsc_nChains        = 4,
                         sparcc_alpha        = 0.05,
                         sparcc_n_boot       = 50,
                         mantel_permutations = 999) {

  # --- Step 1: generate the simulated interaction network ---
  # generateM_specific_type returns a list: [[1]] igraph object, [[2]] GLV matrix M
  simulated_network <- generateM_specific_type(
    nn             = network_size,
    k_ave          = k_average,
    type.network   = "random",
    type.interact  = interaction_type,
    interact.str.max = interact_str_max,
    mix.compt.ratio  = mix_compt_ratio
  )

  true_network <- simulated_network[[1]]
  M            <- simulated_network[[2]]

  # Optionally override diagonal (self-regulation terms); needed for competitive scenarios
  # to prevent the system from diverging
  if (!is.null(self_regulation)) {
    diag(M) <- self_regulation
  }

  # --- Step 2: simulate community data from the GLV model ---
  data_obj <- make_presence_absence_data(
    M,
    n_samples = 900,
    count     = network_size * 10000,
    mode      = 4
  )

  # --- Step 3: remove rare (< 1%) and ubiquitous (> 99%) species ---
  Y <- filterRareCommon(data_obj$pa, lower = 0.01, higher = 0.99)

  # --- Step 4: fit all four models ---
  mrIML_res <- fit_mrIML_network(Y = Y, model_rf = model_rf)
  truth_res <- make_truth_network(M = M, mrIML_mat = mrIML_res$mat)

  m_hmsc <- fit_hmsc_model(
    Y         = Y,
    thin      = hmsc_thin,
    samples   = hmsc_samples,
    transient = hmsc_transient,
    nChains   = hmsc_nChains
  )
  psrf <- check_hmsc_omega_convergence(m_hmsc)

  hmsc_res   <- make_hmsc_ci_network(m_hmsc = m_hmsc, mrIML_mat = mrIML_res$mat,
                                     ci = c(0.025, 0.975), random_level = 1)
  spiec_res  <- fit_spiec_network(data_obj$raw, Y)
  sparcc_res <- fit_sparcc_network(data_obj$raw, Y,
                                   alpha = sparcc_alpha, n_boot = sparcc_n_boot)

  # --- Step 5: visualise ---
  plot_five_networks(mrIML_res, hmsc_res, spiec_res, sparcc_res, truth_res,
                     title_suffix = paste0("(", scenario, ")"))

  # --- Step 6: evaluate recovery ---
  mantel_res <- compare_with_truth(
    true_mat   = truth_res$mat,
    mrIML_mat  = mrIML_res$mat,
    hmsc_mat   = hmsc_res$mat,
    spiec_mat  = spiec_res$mat,
    sparcc_mat = sparcc_res$mat,
    permutations = mantel_permutations
  )

  list(
    scenario     = scenario,
    true_network = true_network,
    M            = M,
    data         = data_obj,
    Y            = Y,
    mrIML        = mrIML_res,
    hmsc_model   = m_hmsc,
    hmsc_psrf    = psrf,
    hmsc         = hmsc_res,
    spiec        = spiec_res,
    sparcc       = sparcc_res,
    truth        = truth_res,
    mantel       = mantel_res
  )
}
```

---

## 3. Network generation helper (`generateM_specific_type`)

The `generateM_specific_type` function (Takemoto et al.) constructs a random GLV interaction matrix `M` with a specified topology and interaction type.

```{r fn-generate-M}
# Generate a GLV interaction matrix with a given network topology and interaction type.
#
# Parameters:
#   nn              – number of species (nodes)
#   k_ave           – average degree (mean edges per node)
#   type.network    – topology: "random" (Erdős-Rényi), "sf" (scale-free),
#                               "sw" (small-world), "bipar" (bipartite)
#   type.interact   – interaction signs: "random", "mutual" (+/+),
#                                        "compt" (-/-), "pp" (+/- predator-prey),
#                                        "mix" (mutualism + competition),
#                                        "mix2" (competition + predator-prey)
#   interact.str.max – maximum absolute interaction strength
#   mix.compt.ratio  – proportion of competitive interactions in "mix"/"mix2"
generateM_specific_type <- function(nn, k_ave,
                                    type.network     = "random",
                                    type.interact    = "random",
                                    interact.str.max = 0.5,
                                    mix.compt.ratio  = 0.5) {
  nl <- round(k_ave * nn / 2)  # number of edges

  # Build the graph topology
  if (type.network == "random") {
    g <- erdos.renyi.game(nn, nl, type = "gnm")
  } else if (type.network == "sf") {
    g <- static.power.law.game(nn, nl, 2.2, -1,
                                loops = FALSE, multiple = FALSE,
                                finite.size.correction = TRUE)
  } else if (type.network == "sw") {
    g <- sample_smallworld(1, nn, round(k_ave / 2), 0.05,
                           loops = FALSE, multiple = FALSE)
  } else if (type.network == "bipar") {
    g <- sample_bipartite(nn / 2, nn / 2, type = "gnm", m = nl, directed = FALSE)
  } else {
    stop("network type is invalid")
  }

  edgelist <- get.edgelist(g)
  A        <- matrix(0, nrow = nn, ncol = nn)

  # Assign interaction strengths based on ecological type
  if (type.interact == "random") {
    for (i in seq_len(nl)) {
      A[edgelist[i, 1], edgelist[i, 2]] <- runif(1, -interact.str.max, interact.str.max)
      A[edgelist[i, 2], edgelist[i, 1]] <- runif(1, -interact.str.max, interact.str.max)
    }
  } else if (type.interact == "mutual") {          # mutualism: both signs positive
    for (i in seq_len(nl)) {
      A[edgelist[i, 1], edgelist[i, 2]] <- runif(1, max = interact.str.max)
      A[edgelist[i, 2], edgelist[i, 1]] <- runif(1, max = interact.str.max)
    }
  } else if (type.interact == "compt") {           # competition: both signs negative
    for (i in seq_len(nl)) {
      A[edgelist[i, 1], edgelist[i, 2]] <- -runif(1, max = interact.str.max)
      A[edgelist[i, 2], edgelist[i, 1]] <- -runif(1, max = interact.str.max)
    }
  } else if (type.interact == "pp") {              # predator-prey: opposing signs
    for (i in seq_len(nl)) {
      s <- if (runif(1) < 0.5) c(1, -1) else c(-1, 1)
      A[edgelist[i, 1], edgelist[i, 2]] <-  s[1] * runif(1, max = interact.str.max)
      A[edgelist[i, 2], edgelist[i, 1]] <-  s[2] * runif(1, max = interact.str.max)
    }
  } else if (type.interact == "mix") {             # mix of mutualism and competition
    for (i in seq_len(nl)) {
      sgn <- if (runif(1) < mix.compt.ratio) -1 else 1
      A[edgelist[i, 1], edgelist[i, 2]] <- sgn * runif(1, max = interact.str.max)
      A[edgelist[i, 2], edgelist[i, 1]] <- sgn * runif(1, max = interact.str.max)
    }
  }

  diag(A) <- -1  # self-regulation (can be overridden via run_scenario)
  list(g, A)
}
```

---

## 4. Model specification

```{r model-spec}
network_size <- 20
k_average    <- 4

# Random forest used by mrIML. mtry and min_n are tuned by cross-validation.
model_rf <- rand_forest(
  trees = 100,
  mode  = "classification",
  mtry  = tune(),
  min_n = tune()
) |>
  set_engine("randomForest")
```

---

## 5. Population dynamics visualisation (optional)

Before running the full benchmarking simulations, it can be helpful to visualise the GLV population dynamics that generate the community data.

```{r dynamics-demo}
# Generate one random network and simulate its dynamics using the GLV model.
# The resulting time-series shows how species abundances fluctuate and interact
# before we sample the final community and convert to presence-absence.
simulated_network_demo <- generateM_specific_type(
  nn               = network_size,
  k_ave            = k_average,
  type.network     = "random",
  type.interact    = "random",
  interact.str.max = 0.4,
  mix.compt.ratio  = 0.5
)

M_demo  <- simulated_network_demo[[2]]
y_demo  <- rpois(network_size, lambda = 100)   # random initial abundances
r_demo  <- runif(network_size)                 # random intrinsic growth rates
res_demo <- glv(network_size, M_demo, r_demo, y_demo)

# Plot species abundances over the last 980 time steps
tsplot(10 * res_demo[, 20:min(1000, ncol(res_demo))], time.given = TRUE)
```

---

## 6. Single-replicate scenarios

Run each of the four ecological scenarios once, producing network visualisations and Mantel statistics.

```{r single-reps}
random_res <- run_scenario(
  scenario         = "Random",
  interaction_type = "random",
  interact_str_max = 0.4,
  network_size     = network_size,
  k_average        = k_average,
  mix_compt_ratio  = 0.5,
  model_rf         = model_rf
)

mutual_res <- run_scenario(
  scenario         = "Mutualistic",
  interaction_type = "mutual",
  interact_str_max = 0.3,
  network_size     = network_size,
  k_average        = k_average,
  mix_compt_ratio  = 0.5,
  model_rf         = model_rf
)

competitive_res <- run_scenario(
  scenario         = "Competitive",
  interaction_type = "compt",
  interact_str_max = 0.5,
  network_size     = network_size,
  k_average        = k_average,
  mix_compt_ratio  = 0.5,
  self_regulation  = -0.5,   # stronger self-regulation prevents divergence under pure competition
  model_rf         = model_rf
)

mixed_res <- run_scenario(
  scenario         = "Mixed",
  interaction_type = "mix",
  interact_str_max = 0.4,
  network_size     = network_size,
  k_average        = k_average,
  mix_compt_ratio  = 0.5,
  model_rf         = model_rf
)
```

### Summary table (single replicate)

```{r single-table}
scenarios <- list(random_res, mutual_res, competitive_res, mixed_res)

comparison_table <- tibble(
  Scenario = c("Random", "Mutualistic", "Competitive", "Mixed"),
  mrIML_r  = round(sapply(scenarios, \(s) s$mantel$mrIML$statistic),      3),
  mrIML_p  =       sapply(scenarios, \(s) s$mantel$mrIML$signif),
  Hmsc_r   = round(sapply(scenarios, \(s) s$mantel$Hmsc$statistic),       3),
  Hmsc_p   =       sapply(scenarios, \(s) s$mantel$Hmsc$signif),
  SPIEC_r  = round(sapply(scenarios, \(s) s$mantel$SPIEC_EASI$statistic), 3),
  SPIEC_p  =       sapply(scenarios, \(s) s$mantel$SPIEC_EASI$signif),
  SparCC_r = round(sapply(scenarios, \(s) s$mantel$SparCC$statistic),     3),
  SparCC_p =       sapply(scenarios, \(s) s$mantel$SparCC$signif)
)

DT::datatable(
  comparison_table,
  colnames = c(
    "Scenario",
    "mrIML r", "mrIML p",
    "Hmsc 95% CrI r", "Hmsc 95% CrI p",
    "SPIEC-EASI r", "SPIEC-EASI p",
    "SparCC r", "SparCC p"
  ),
  options = list(pageLength = 10, dom = "t")
)
```

---

## 7. Replicated simulations

A single replicate is sensitive to the random network draw. The `run_scenarios_replicated()` wrapper repeats each scenario `n_reps` times, accumulating Mantel statistics and summarising them as means with 95% quantile confidence intervals.

**Parallelism note:** mrIML saturates available cores within each replicate (one worker per species model), so the outer replicate loop is kept sequential to avoid resource contention.

**Checkpointing:** Each completed replicate is saved to disk. If the session crashes, re-running the same call will skip already-completed replicates automatically.

```{r replicated-setup}
# Load the replicated-scenario wrapper and default scenario configurations
source("generateM_specific_type.R")
source("run_scenarios_replicated.R")

library(future)
library(furrr)

plan(multisession, workers = parallel::detectCores() - 2)
```

```{r replicated-run}
# Run each scenario n_reps = 5 times.
# Reduce hmsc_* settings for a quick smoke test; use defaults for final results.
rep_results <- run_scenarios_replicated(
  scenarios_config    = default_scenarios_config,
  n_reps              = 5,
  model_rf            = model_rf,
  mantel_permutations = 99,
  hmsc_thin           = 10,
  hmsc_samples        = 500,
  hmsc_transient      = 2000,
  hmsc_nChains        = 2,
  checkpoint_dir      = "sim_checkpoints"  # saved per-rep RDS files go here
)
```

### Inspect results

```{r replicated-results}
# Long table: one row per scenario × rep × model
rep_results$raw

# Summary: mean Mantel r [95% quantile CI], mean p [95% CI], proportion significant
rep_results$summary

# Wide format: one row per scenario, all models as columns
rep_results$wide

# Interactive DT table with colour-coded significance bars
pretty_ci_table(rep_results$summary)
```

### Publication-ready table (Word)

```{r pub-table}
library(flextable)
library(officer)

pub_table <- rep_results$summary |>
  select(scenario, model, n_reps_ok, r_95CI, p_95CI, prop_sig_p) |>
  rename(
    Scenario                = scenario,
    Model                   = model,
    `Reps (OK)`             = n_reps_ok,
    `r (mean [95% CI])`     = r_95CI,
    `p (mean [95% CI])`     = p_95CI,
    `Prop. sig. (p < 0.05)` = prop_sig_p
  )

ft <- flextable(pub_table) |>
  merge_v(j = "Scenario") |>               # merge repeated scenario labels
  valign(j = "Scenario", valign = "top") |>
  theme_booktabs() |>
  bold(part = "header") |>
  autofit() |>
  set_caption(paste(
    "Table 1. Summary of co-occurrence model performance across simulation scenarios.",
    "r: Spearman Mantel correlation between predicted and true co-occurrence network;",
    "p: associated p-value;",
    "Prop. sig.: proportion of replicates with p < 0.05."
  ))

doc <- read_docx() |>
  body_add_flextable(ft)

print(doc, target = "results_summary_table.docx")
```

---

## 8. `run_scenarios_replicated()` internals

This section documents the helper functions sourced from `run_scenarios_replicated.R`.

### Extract Mantel statistics

```{r fn-extract-mantel}
# Return a small data frame of r and p for all four models in one replicate.
.extract_mantel_df <- function(mantel_list, scenario, rep) {
  models <- c("mrIML", "Hmsc", "SPIEC_EASI", "SparCC")
  data.frame(
    scenario = scenario,
    rep      = rep,
    model    = models,
    r        = sapply(models, \(m) mantel_list[[m]]$statistic),
    p        = sapply(models, \(m) mantel_list[[m]]$signif),
    error    = NA_character_,
    row.names = NULL, stringsAsFactors = FALSE
  )
}
```

### Run one replicate with error handling and checkpointing

```{r fn-run-one-rep}
# Execute a single scenario × rep, wrapping run_scenario() in tryCatch so
# that one failed replicate does not abort the whole simulation.
# Plots are redirected to a null device to avoid opening dozens of windows.
# Heavy objects (Hmsc MCMC chains, mrIML bootstraps, raw count data) are
# explicitly freed before returning to keep peak memory low.
.run_one_rep <- function(sc_name, sc_args, rep, model_rf, mantel_permutations,
                         checkpoint_dir, seed) {

  checkpoint_file <- if (!is.null(checkpoint_dir)) {
    file.path(checkpoint_dir, paste0(sc_name, "_rep", rep, ".rds"))
  } else NULL

  # Return cached result without re-running
  if (!is.null(checkpoint_file) && file.exists(checkpoint_file)) {
    message("  [cached] ", sc_name, " rep ", rep)
    return(readRDS(checkpoint_file))
  }

  call_args <- c(
    list(scenario            = paste0(sc_name, "_rep", rep),
         model_rf            = model_rf,
         mantel_permutations = mantel_permutations),
    sc_args
  )

  result <- tryCatch({
    grDevices::pdf(file = nullfile())     # suppress plot output
    null_dev <- grDevices::dev.cur()
    on.exit(grDevices::dev.off(null_dev), add = TRUE)

    res <- do.call(run_scenario, call_args)
    out <- .extract_mantel_df(res$mantel, sc_name, rep)

    rm(res); gc(verbose = FALSE, full = TRUE)  # free large objects immediately
    out
  },
  error = function(e) {
    msg <- conditionMessage(e)
    message("ERROR — Scenario '", sc_name, "' rep ", rep, ": ", msg)
    models <- c("mrIML", "Hmsc", "SPIEC_EASI", "SparCC")
    data.frame(scenario = sc_name, rep = rep, model = models,
               r = NA_real_, p = NA_real_, error = msg,
               row.names = NULL, stringsAsFactors = FALSE)
  })

  # Only cache successful results; skip caching failures so they can be retried
  if (!is.null(checkpoint_file) && !any(is.na(result$r))) {
    saveRDS(result, checkpoint_file)
  }

  result
}
```

### Main wrapper

```{r fn-run-scenarios-replicated}
# Main entry point. Loops over all scenario × rep combinations sequentially,
# then summarises Mantel r and p with quantile confidence intervals.
run_scenarios_replicated <- function(scenarios_config,
                                     n_reps,
                                     model_rf,
                                     ci_alpha            = 0.05,
                                     sig_alpha           = 0.05,
                                     seed                = NULL,
                                     mantel_permutations = 499,
                                     hmsc_thin           = 50,
                                     hmsc_samples        = 1000,
                                     hmsc_transient      = 5000,
                                     hmsc_nChains        = 4,
                                     checkpoint_dir      = "sim_checkpoints",
                                     verbose             = TRUE) {

  stopifnot(is.list(scenarios_config), length(scenarios_config) > 0)
  stopifnot(n_reps >= 2)

  if (!is.null(checkpoint_dir) && !dir.exists(checkpoint_dir)) {
    dir.create(checkpoint_dir, recursive = TRUE)
    if (verbose) message("Created checkpoint directory: ", checkpoint_dir)
  }

  n_jobs   <- length(scenarios_config) * n_reps
  raw_list <- vector("list", n_jobs)
  idx      <- 1L

  for (sc_name in names(scenarios_config)) {
    for (rep in seq_len(n_reps)) {
      if (verbose) message("[", idx, "/", n_jobs, "] ", sc_name, " rep ", rep)
      if (!is.null(seed)) set.seed(seed + idx - 1L)

      raw_list[[idx]] <- .run_one_rep(
        sc_name             = sc_name,
        sc_args             = c(scenarios_config[[sc_name]],
                                list(hmsc_thin      = hmsc_thin,
                                     hmsc_samples   = hmsc_samples,
                                     hmsc_transient = hmsc_transient,
                                     hmsc_nChains   = hmsc_nChains)),
        rep                 = rep,
        model_rf            = model_rf,
        mantel_permutations = mantel_permutations,
        checkpoint_dir      = checkpoint_dir,
        seed                = NULL
      )
      idx <- idx + 1L
    }
  }

  raw <- do.call(rbind, raw_list)

  # Summarise: quantile CI for r and p, proportion of replicates with p < sig_alpha
  probs_lo <- ci_alpha / 2
  probs_hi <- 1 - ci_alpha / 2

  summary_tbl <- raw |>
    dplyr::group_by(scenario, model) |>
    dplyr::summarise(
      n_reps_ok  = sum(!is.na(r)),
      mean_r     = mean(r,  na.rm = TRUE),
      ci_lo_r    = quantile(r, probs_lo, na.rm = TRUE),
      ci_hi_r    = quantile(r, probs_hi, na.rm = TRUE),
      mean_p     = mean(p,  na.rm = TRUE),
      ci_lo_p    = quantile(p, probs_lo, na.rm = TRUE),
      ci_hi_p    = quantile(p, probs_hi, na.rm = TRUE),
      prop_sig_p = mean(p < sig_alpha, na.rm = TRUE),
      .groups    = "drop"
    ) |>
    dplyr::mutate(
      dplyr::across(c(mean_r, ci_lo_r, ci_hi_r, mean_p, ci_lo_p, ci_hi_p),
                    \(x) round(x, 3)),
      prop_sig_p = round(prop_sig_p, 2),
      r_95CI = paste0(mean_r, " [", ci_lo_r, ", ", ci_hi_r, "]"),
      p_95CI = paste0(mean_p, " [", ci_lo_p, ", ", ci_hi_p, "]")
    )

  wide_tbl <- summary_tbl |>
    dplyr::select(scenario, model, r_95CI, p_95CI, prop_sig_p) |>
    tidyr::pivot_wider(
      names_from  = model,
      values_from = c(r_95CI, p_95CI, prop_sig_p),
      names_glue  = "{model}_{.value}"
    )

  if (verbose) message("Done.")
  list(raw = raw, summary = summary_tbl, wide = wide_tbl)
}
```

### Default scenario configurations

```{r default-configs}
# Four scenarios matching the original supplementary page.
# Each list element is passed as extra arguments to run_scenario().
default_scenarios_config <- list(
  Random = list(
    interaction_type = "random",
    interact_str_max = 0.4,
    mix_compt_ratio  = 0.5
  ),
  Mutualistic = list(
    interaction_type = "mutual",
    interact_str_max = 0.3,
    mix_compt_ratio  = 0.5
  ),
  Competitive = list(
    interaction_type = "compt",
    interact_str_max = 0.5,
    mix_compt_ratio  = 0.5,
    self_regulation  = -0.5
  ),
  Mixed = list(
    interaction_type = "mix",
    interact_str_max = 0.4,
    mix_compt_ratio  = 0.5
  )
)
```

---

## Notes on MCMC settings

For final published results, use the longer Hmsc MCMC settings:

```{r hmsc-long, eval=FALSE}
random_res_long <- run_scenario(
  scenario         = "Random - longer Hmsc",
  interaction_type = "random",
  interact_str_max = 0.4,
  network_size     = network_size,
  k_average        = k_average,
  mix_compt_ratio  = 0.5,
  model_rf         = model_rf,
  hmsc_thin        = 100,
  hmsc_samples     = 2000,
  hmsc_transient   = 10000,
  hmsc_nChains     = 4
)
```

Check convergence diagnostics (Gelman-Rubin R-hat) before interpreting Hmsc results — values consistently below 1.1 across all Omega parameters suggest adequate mixing.
