---
title: "Case study: Forest communities (beetles, birds and plants)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case study: Forest communities (beetles, birds and plants)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  eval    = FALSE,
  warning = FALSE,
  message = FALSE
)
```

> This case study is adapted from the mrIML 2.0 supplementary material authored by Ryan K. Leadbetter and Nicholas M. Fountain-Jones ([source](https://rleadbett.github.io/mrIML-paper-supmat/case-studies/02-forest-communities/dummy.html)).

## Overview

This case study applies mrIML 2.0 to a multi-taxonomic dataset spanning beetles, birds, and plants across a forest successional gradient in Tasmania, Australia. The dataset comprises 144 plots representing four forest age categories:

- **Mature forest** — unharvested with no fire disturbance for ≥ 60 years
- **~45 years post-harvest**
- **~27 years post-harvest**
- **~7 years post-harvest**

Environmental covariates include coarse wood debris, leaf-litter depth, and soil variables (pH, conductivity). For each taxonomic group, three model configurations are compared — environment only, co-occurrence only, and combined — to assess the relative contributions of abiotic filtering and biotic interactions to community assembly.

**Key findings:** Environmental filtering dominates across all groups. Plant communities show the strongest responses to successional gradients, while bird and beetle communities respond primarily to altitude and microclimate.

```{r libraries}
library(here)
library(readxl)
library(tidyverse)
library(DT)
library(mrIML)
library(tidymodels)
library(future)
library(finetune)
library(flashlight)
library(igraph)
library(ggnetwork)
library(cowplot)
library(patchwork)

set.seed(123)
n_cores <- parallel::detectCores()
plan("multisession", workers = n_cores - 2)

# Path to data files (adjust to your local directory)
data_dir <- here("case-studies", "02-forest-communities", "data", "raw")
```

## Data preparation

### Environmental covariates

```{r predictors}
raw_predictors <- read_excel(
  file.path(data_dir, "Site_covariates_decision_making_4.xlsx"),
  sheet = 4
) |>
  column_to_rownames(var = "plot_code") |>
  rename_all(~ gsub(" ", "_", .)) |>
  dplyr::select(-plot_code_T_D, -Transect)

datatable(raw_predictors)
```

### Species matrices

```{r species-data}
raw_data_beetles <- read_excel(
  file.path(data_dir, "all_taxa_all_sp_mat_sp.xlsx"), sheet = "beetles_all"
) |>
  dplyr::select(-Site, -Distance, -Transect, -Age) |>
  rename_all(~ gsub(" ", "_", .)) |>
  column_to_rownames(var = "Plot") |>
  mutate_all(~ ifelse(. > 0, 1, .))   # convert abundance to presence/absence

raw_data_birds <- read_excel(
  file.path(data_dir, "all_taxa_all_sp_mat_sp.xlsx"), sheet = "birds_all"
) |>
  dplyr::select(-Site, -Distance, -Transect, -Age) |>
  rename_all(~ gsub(" ", "_", .)) |>
  column_to_rownames(var = "Plot") |>
  mutate_all(~ ifelse(. > 0, 1, .))

raw_data_plants <- read_excel(
  file.path(data_dir, "all_taxa_all_sp_mat_sp.xlsx"), sheet = "plants_all"
) |>
  dplyr::select(-Site, -Distance, -Transect, -Age) |>
  rename_all(~ gsub(" ", "_", .)) |>
  column_to_rownames(var = "Plot") |>
  mutate_all(~ ifelse(. > 0, 1, .))
```

## Model specification

All three taxonomic groups use the same random-forest base model.

```{r model-spec}
model_rf <- rand_forest(
  trees = 1000,
  mode  = "classification",
  mtry  = tune(),
  min_n = tune()
) |>
  set_engine("randomForest")
```

---

## Beetle community

### Prepare X and Y

Species with prevalence outside 20–70% are filtered out to avoid near-constant responses. Environmental variables are selected via an internal selection helper before modelling.

```{r beetle-data}
# Retain species with 20–70% prevalence
Y_beetles <- filterRareCommon(raw_data_beetles, lower = 0.2, higher = 0.7)

# Remove beetle- and bird-specific columns not relevant as general predictors
X_beetles <- raw_predictors |>
  dplyr::select(
    -Beetles_S_Mat_Indval2, -Beetle_S_Total, -dist_regrowth, -Age,
    -Beetles_Mat_Ratio, -Birds_S_Mat_EFL, -Birds_S_Total, -Birds_Mat_Ratio,
    -slope_f_Jayne, -aspect_f_Jayne
  )

X_beetles$Site     <- as.factor(X_beetles$Site)
X_beetles$geol_Jayne <- as.factor(X_beetles$geol_Jayne)

# Variable selection (removes low-importance predictors)
var_sel_beetles <- select_variables(Y = Y_beetles, X = X_beetles)
X_beetles <- var_sel_beetles[[2]]
X_beetles$Site <- NULL   # remove site to enable clean model comparison

X1_beetles <- Y_beetles  # co-occurrence predictors = all beetle species
```

### Fit three models

```{r beetle-fit}
# Environment only
yhats_beetles_env <- mrIMLpredicts(
  X = X_beetles, Y = Y_beetles,
  Model = model_rf, balance_data = "no",
  tune_grid_size = 5, prop = 0.6, k = 5, racing = TRUE
)

# Co-occurrence only
yhats_beetles_cooccur <- mrIMLpredicts(
  Y = Y_beetles, X1 = X1_beetles,
  Model = model_rf, balance_data = "no",
  tune_grid_size = 5, prop = 0.6, k = 5, racing = TRUE
)

# Combined
yhats_beetles_combined <- mrIMLpredicts(
  X = X_beetles, Y = Y_beetles, X1 = X1_beetles,
  Model = model_rf, balance_data = "no",
  tune_grid_size = 5, prop = 0.6, k = 5, racing = TRUE
)
```

### Performance

```{r beetle-performance}
ModelPerf_beetles_env      <- mrIMLperformance(yhats_beetles_env)
ModelPerf_beetles_cooccur  <- mrIMLperformance(yhats_beetles_cooccur)
ModelPerf_beetles_combined <- mrIMLperformance(yhats_beetles_combined)

# Combined vs. environment-only comparison plot
perf_comparison_beetles <- mrPerformancePlot(
  ModelPerf1 = ModelPerf_beetles_combined,
  ModelPerf2 = ModelPerf_beetles_env,
  mode       = "classification"
)
perf_comparison_beetles$performance_plot +
  labs(title = "Beetle community model performance")

# MCC summary table
data.frame(
  Model = c("Environment only", "Co-occurrence only", "Combined"),
  MCC   = c(
    ModelPerf_beetles_env[[2]],
    ModelPerf_beetles_cooccur[[2]],
    ModelPerf_beetles_combined[[2]]
  )
)
```

### Variable importance

```{r beetle-vip}
bs_beetles <- mrBootstrap(yhats_beetles_combined, num_bootstrap = 100)

vi_beetles <- mrVip(
  yhats_beetles_combined,
  mrBootstrap_obj = bs_beetles,
  threshold       = 0.5,
  global_top_var  = 10
)

vi_beetles[[3]] + labs(title = "Beetle community variable importance")
```

### Co-occurrence network

```{r beetle-network}
assoc_net_beetles <- mrCoOccurNet(bs_beetles)
mrCoOccurNet_plot(
  assoc_net_beetles,
  strength_thresh = 0.1,
  network_title   = "Beetle co-occurrence network"
)
```

### Interactions

```{r beetle-interactions}
interactions_rf <- yhats_beetles_combined |>
  mrInteractions(num_bootstrap = 10)

interactions_rf[[1]]
```

---

## Bird community

The bird analysis mirrors the beetle analysis structure. The same environmental predictors are used, with variable selection applied separately.

```{r bird-data}
Y_birds <- filterRareCommon(raw_data_birds, lower = 0.2, higher = 0.7)

X_birds <- raw_predictors |>
  dplyr::select(
    -Beetles_S_Mat_Indval2, -Beetle_S_Total, -dist_regrowth, -Age,
    -Beetles_Mat_Ratio, -Birds_S_Mat_EFL, -Birds_S_Total, -Birds_Mat_Ratio,
    -slope_f_Jayne, -aspect_f_Jayne
  )

X_birds$Site     <- as.factor(X_birds$Site)
X_birds$geol_Jayne <- as.factor(X_birds$geol_Jayne)
X_birds$Site     <- NULL

var_sel_birds <- select_variables(Y = Y_birds, X = X_birds)
X_birds  <- var_sel_birds[[2]]
X1_birds <- Y_birds
```

```{r bird-fit}
yhats_birds_env <- mrIMLpredicts(
  X = X_birds, Y = Y_birds,
  Model = model_rf, balance_data = "no",
  tune_grid_size = 5, prop = 0.6, k = 5, racing = TRUE
)

yhats_birds_cooccur <- mrIMLpredicts(
  Y = Y_birds, X1 = X1_birds,
  Model = model_rf, balance_data = "no",
  tune_grid_size = 5, prop = 0.6, k = 5, racing = TRUE
)

yhats_birds_combined <- mrIMLpredicts(
  X = X_birds, Y = Y_birds, X1 = X1_birds,
  Model = model_rf, balance_data = "no",
  tune_grid_size = 5, prop = 0.6, k = 5, racing = TRUE
)
```

```{r bird-performance}
ModelPerf_birds_env      <- mrIMLperformance(yhats_birds_env)
ModelPerf_birds_cooccur  <- mrIMLperformance(yhats_birds_cooccur)
ModelPerf_birds_combined <- mrIMLperformance(yhats_birds_combined)

perf_comparison_birds <- mrPerformancePlot(
  ModelPerf1 = ModelPerf_birds_combined,
  ModelPerf2 = ModelPerf_birds_env,
  mode       = "classification"
)
perf_comparison_birds$performance_plot + labs(title = "Bird community model performance")

data.frame(
  Model = c("Environment only", "Co-occurrence only", "Combined"),
  MCC   = c(
    ModelPerf_birds_env[[2]],
    ModelPerf_birds_cooccur[[2]],
    ModelPerf_birds_combined[[2]]
  )
)
```

```{r bird-vip}
bs_birds <- mrBootstrap(yhats_birds_combined, num_bootstrap = 100)

vi_birds <- mrVip(
  yhats_birds_combined,
  mrBootstrap_obj = bs_birds,
  threshold       = 0.5,
  global_top_var  = 10
)

vi_birds[[3]] + labs(title = "Bird community variable importance")
```

```{r bird-network}
assoc_net_birds <- mrCoOccurNet(bs_birds)
mrCoOccurNet_plot(
  assoc_net_birds,
  strength_thresh = 0.1,
  network_title   = "Bird co-occurrence network"
)
```

---

## Plant community

Plants show the strongest response to the successional gradient. Additional predictor columns specific to vegetation surveys are removed before modelling.

```{r plant-data}
Y_plants <- filterRareCommon(raw_data_plants, lower = 0.2, higher = 0.7)

X_plants <- raw_predictors |>
  dplyr::select(
    -Beetles_S_Mat_Indval2, -Beetle_S_Total, -dist_regrowth, -Age,
    -Beetles_Mat_Ratio, -Birds_S_Mat_EFL, -Birds_S_Total, -Birds_Mat_Ratio,
    -slope_f_Jayne, -aspect_f_Jayne,
    -contains("Bryo"), -contains("Plant"), -contains("Moss"),
    -Vegetation_cover
  ) |>
  rename_with(~ gsub("Jayne", "", .), contains("Jayne"))

X_plants$Site  <- as.factor(X_plants$Site)
X_plants$geol_ <- as.factor(X_plants$geol_)
X_plants$Site  <- NULL

var_sel_plants <- select_variables(Y = Y_plants, X = X_plants)
X_plants  <- var_sel_plants[[2]]
X1_plants <- Y_plants
```

```{r plant-fit}
yhats_plants_env <- mrIMLpredicts(
  X = X_plants, Y = Y_plants,
  Model = model_rf, balance_data = "no",
  tune_grid_size = 5, prop = 0.6, k = 5, racing = TRUE
)

yhats_plants_cooccur <- mrIMLpredicts(
  Y = Y_plants, X1 = X1_plants,
  Model = model_rf, balance_data = "no",
  tune_grid_size = 5, prop = 0.6, k = 5, racing = TRUE
)

yhats_plants_combined <- mrIMLpredicts(
  X = X_plants, Y = Y_plants, X1 = X1_plants,
  Model = model_rf, balance_data = "no",
  tune_grid_size = 5, prop = 0.6, k = 5, racing = TRUE
)
```

```{r plant-performance}
ModelPerf_plants_env      <- mrIMLperformance(yhats_plants_env)
ModelPerf_plants_cooccur  <- mrIMLperformance(yhats_plants_cooccur)
ModelPerf_plants_combined <- mrIMLperformance(yhats_plants_combined)

perf_comparison_plants <- mrPerformancePlot(
  ModelPerf1 = ModelPerf_plants_combined,
  ModelPerf2 = ModelPerf_plants_env,
  mode       = "classification"
)
perf_comparison_plants$performance_plot + labs(title = "Plant community model performance")

data.frame(
  Model = c("Environment only", "Co-occurrence only", "Combined"),
  MCC   = c(
    ModelPerf_plants_env[[2]],
    ModelPerf_plants_cooccur[[2]],
    ModelPerf_plants_combined[[2]]
  )
)
```

```{r plant-vip}
bs_plants <- mrBootstrap(yhats_plants_combined, num_bootstrap = 100)

vi_plants <- mrVip(
  yhats_plants_combined,
  mrBootstrap_obj = bs_plants,
  threshold       = 0.5,
  global_top_var  = 10
)

vi_plants[[3]] + labs(title = "Plant community variable importance")
```

```{r plant-network}
assoc_net_plants <- mrCoOccurNet(bs_plants)
mrCoOccurNet_plot(
  assoc_net_plants,
  strength_thresh = 0.1,
  network_title   = "Plant co-occurrence network",
  nmds_layout     = TRUE
)
```
