---
title: "Graphical network model (GNM)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{`Graphical network model (GNM)`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
    fig.width = 8, fig.height = 6, fig.align = "center",
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

This vignette guides you through the new functionality that turns a `mrIML` multi-response model into a graphical network model (GNM). There are also new functions in `mrIML (>= 2.0.0)` that can be used in non-GNM models, such as for single-taxon SDMs, epi models, gene-environment association (GEA) studies, and landscape genetics. Advances in [Tidymodels](https://www.tidymodels.org) are at the core of these functions, as are exciting developments in interpretable machine learning (e.g., for quickly quantifying interactions through H-statistics). The capability to capture uncertainty via bootstraps and <span style="color:red">spatial patterns</span> are also features of this `mrIML` module. The examples come from the microbiome world but are, of course, generalizable to any multi-response problem.

Currently, the models are optimized and tested on presence/absence data. Models suitable for abundance data will be added in the future.

To demonstrate, first, we need to load some packages and data. The first example uses coinfection data from New Caledonian *Zosterops* species. A single continuous covariate is also included (`scale.prop.zos`), which reflects the relative abundance of *Zosterops* species among different sample sites.

```{r libraries}
library(mrIML)
library(tidymodels)
library(flashlight)
library(ggplot2)
library(patchwork)
library(igraph)
library(ggnetwork)
library(network)

set.seed(7007)
```

```{r load_data}
data <- MRFcov::Bird.parasites
Y <-  data %>%
  dplyr::select(-scale.prop.zos) %>%
  dplyr::select(order(everything()))

X <- data %>%
  dplyr::select(
    scale.prop.zos
  )

X1 <- Y
```

Note that the inclusion of `X1` converts `mrIML` into a Joint Species Distribution Model (JSDM) by directly adding the presence/absence patterns of the other taxa into the model as predictors (along with other environmental/host covariates). Note that the order of `X1` and `Y` needs to match.

## Setting up the models

Built into the `mrIML` architecture (and a big advantage of `tidymodels`) is the capability to change the underlying model easily. We are going to set up two models to compare: a random forest model (RF) and logistic regression (LM). `mrIML` takes advantage of multi-core processing, so we set it up here to run on 5 cores. These steps are the same as in `mrIML (< 2.0.0)`.

```{r parallel_set_up, eval=FALSE}
future::plan("multisession", workers = 5)
```

```{r define_models}
model_rf <-rand_forest(
  trees = 100,
  mode = "classification",
  mtry = tune(),
  min_n = tune()
) %>%
  set_engine("randomForest")

model_lm <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")
```

## Running the models

Aside from adding JSDM functionality with the `X1` call, we have also enabled `mrIML` to tune hyperparameters using the very efficient `racing` option (see Kuhn (2014)). In brief, this racing option takes a small subsample of parameters and eliminates combinations that do not improve fit using a repeated-measures ANOVA model with [`finetune::tune_race_anova()`](https://finetune.tidymodels.org/reference/tune_race_anova.html). Setting `racing = FALSE` reverts to a grid search if you want to manually set the tuning grid size (i.e., using [`tune::tune_grid()`](https://tune.tidymodels.org/reference/tune_grid.html)). For logistic regression, there are no parameters to tune, so set `racing = FALSE`.

```{r fit_mrIML_models}
#random forest
yhats_rf <- mrIMLpredicts(
  X = X,
  Y = Y,
  X1 = Y,
  Model = model_rf,
  prop = 0.7,
  k = 5,
  racing = TRUE
)

#linear model
yhats_lm <- mrIMLpredicts(
  X = X,
  Y = Y,
  X1 = Y,
  Model = model_lm , 
  balance_data = "no",
  prop = 0.6,
  k = 5,
  racing = FALSE
)
```

## Comparing performance

It's important to compare whether there are any advantages to using a random forest approach. Interpretation would be easier overall if logistic regression gave similar predictive performance.

```{r perf}
ModelPerf_rf <- mrIMLperformance(yhats_rf)

ModelPerf_rf[[1]] #across all parasites
ModelPerf_rf[[2]] #overall

ModelPerf_lm <- mrIMLperformance(yhats_lm)

ModelPerf_lm[[1]]
ModelPerf_lm[[2]]

plots <- mrPerformancePlot(
  ModelPerf1 = ModelPerf_lm,
  ModelPerf2 = ModelPerf_rf,
  mode = "classification"
)

plots[[1]] /
plots[[2]]
plots[[3]]
```

If we just look at the overall AUC values, model performance appears quite similar (`r mean(ModelPerf_rf[[1]]$roc_AUC) %>% round(digits = 2)` for the RF and `r mean(ModelPerf_lm[[1]]$roc_AUC) %>% round(digits = 2)` for the LM). However, when we look at the Matthews correlation coefficient (MCC) for each taxon in the LM model, we see that for *H.killangoi* and Plas (*Plasmodium*), the values are much lower (an overall MCC of `r mean(ModelPerf_lm[[2]]) %>% round(digits = 2)` for LM is basically just a guess, compared to `r mean(ModelPerf_rf[[2]]) %>% round(digits = 2)` for the RF). Remember, the classes are imbalanced, so AUC tends to be an inflated measure. This provides evidence that non-linear relationships can improve predictive performance.

Next, we ask whether including putative associations between taxa improves model performance, or whether the relationship between parasites and host relative abundance alone is sufficient (i.e., assuming independence between the parasites).

```{r independent_responses}
yhats_rf_noAssoc <- mrIMLpredicts(
  X = X,
  Y = Y,
  Model = model_rf,
  prop = 0.7,
  k = 5,
  racing = TRUE
)

ModelPerf_rf_noAssoc <- mrIMLperformance(yhats_rf_noAssoc)

ModelPerf_rf_noAssoc[[1]]
ModelPerf_rf[[1]] #performance including associations
```

You can see that overall, including associations improves model performance, particularly for predicting *H.killangoi* and *Microfilaria*. Using MCC to compare models is problematic because, in the association-free model, it's undefined for these taxa (`NA` values arise because there are no false negatives for low-prevalence taxa). <span style="color:red">Positive predictive value (PPV) is useful in this case and shows that without associations, we can't predict the occurrence of these taxa (PPV = 0 for both). Including associations increases PPV to \~0.2—not great, as 80% of our positive predictions for these taxa are false.</span> 

## Downsampling

Including associations makes a difference, but how can we improve predictions for our two rarer taxa? Upsampling is possible, but in this case, we'll try downsampling to see if correcting class imbalance improves model fit.

```{r downsampling}
yhats_rf_downSamp <- mrIMLpredicts(
  X = X,
  Y = Y,
  X1 = Y,
  Model = model_rf,
  balance_data = "down", #down sampling
  prop = 0.75,
  k = 5,
  racing = TRUE)

ModelPerf_rf_downSamp <- mrIMLperformance(yhats_rf_downSamp)
ModelPerf_rf_downSamp[[1]]
```

Look at those PPV values now—much better. <span style="color:red">Our false positive rate is down to <6% overall. Now that we are happy with the performance of the model, we can interrogate it further.</span>

## Interpreting the model

In many cases, like this dataset, community or microbiome data tend to be small in size. Applying stochastic machine learning algorithms to such data can lead to challenges. For instance, variable importance may vary substantially when we build multiple models using the same data and algorithm. To handle this variability and better understand prediction uncertainty, `mrIML` now includes functionality to capture uncertainty in the tuned model using bootstrapping. <span style="color:red">Additionally, this approach helps estimate how variables affect the response, and these estimates align with those from traditional linear regression models (see Cook et al., 2021).</span>

`mrIML` makes it easy to obtain bootstrap estimates for a variety of interpretable machine learning tools and uses these estimates to construct marginalized co-occurrence networks. First, let's perform bootstrapping and calculate variable importance.

```{r bootstraps}
bs_malaria <- mrBootstrap(
  yhats_rf,
  num_bootstrap = 100,
  downsample = TRUE
)

bs_impVI <- mrVip(
  mrIMLobj = yhats_rf,
  mrBootstrap_obj = bs_malaria,
  model_perf = ModelPerf_rf
)

bs_impVI[[3]]
```

You can see that host abundance is the most important predictor of this parasite community (followed by *H.zosteropsis*). However, the sub-figure on the right shows substantial variability. For example, *H.zosteropsis* is the most important predictor for the occurrence of *Microfilaria*, and host abundance (shortened to "sc..") is less important.

## Bootstrap partial dependence plots

To examine the relationship between each variable and community structure, `mrIML` has a convenient wrapper to plot bootstrapped partial dependencies for a taxon of interest.

```{r bootstrapped_PDs}
pds <- mrPdPlotBootstrap(
  yhats_rf,
  mrBootstrap_obj=bs_malaria,
  vi_obj=bs_impVI,
  target='Plas',
  global_top_var=5
)
pds[[2]]
```

These plots show that the presence of *Microfilaria* greatly increases the probability of observing *Plasmodium* (from \~0.47 to 0.70 while holding all other variables at their mean value). Note that these are marginal relationships (i.e., isolating the effect of each predictor). When host abundance is high, the probability of detecting *Plasmodium* decreases non-linearly (with a threshold around \~0.5). The other parasites have less effect.

To explore the effect of host abundance overall, we can use the `mrCovar()` function.

```{r host_abundance_effect}
covar <- mrCovar(
  yhats_rf,
  var = "scale.prop.zos",
  sdthresh = 0.01
)

covar[[1]] /
  covar[[2]] /
  covar[[3]]
```

Note that this isn't bootstrapped—each line represents a taxon. The second plot shows the community-wide average effect with taxon-specific effects silhouetted in the background. The third plot shows the community-wide average change in occurrence probabilities across host abundance. <span style="color:red">Note that the occurrence probabilities of all taxa drop at intermediate levels of host abundance (0.75–1.25).</span>

## Co-occurrence network

We can utilize all the bootstrapped partial dependence estimates (PDs) to construct a co-occurrence network. The following code converts the object to an `igraph` object and plots it. This is a directed network, and edges are scaled by the standard deviation of the marginal change in prediction (along the PD curves). Red edges are positive associations (i.e., predicted occurrence increases with the presence of the other taxon), and blue edges are negative.

```{r network_plot}
assoc_net<- mrCoOccurNet(bs_malaria)

# Based on simulations (not shown here), we use 
# the following rule of thumb for associations:
# Any association > 0.05 for mean strength is 
# included
assoc_net_filtered <-  assoc_net %>% 
  filter(mean_strength > 0.05)

# Convert mrIML to igraph
g <- graph_from_data_frame(
  assoc_net_filtered,
  directed = TRUE,
  vertices = names(Y)
)

E(g)$Value <- assoc_net_filtered$mean_strength

E(g)$Color <- ifelse(
  assoc_net_filtered$direction == "negative",
  "blue",
  "red"
)

# Convert the igraph to a ggplot with NMDS layout
gg <- ggnetwork(g)

# Plot the graph
gg %>%
  ggplot(
    aes(
      x = x, y = y,
      xend = xend, yend = yend
    )
  ) +
  geom_edges(
    aes(color = Color, linewidth = Value), 
    curvature = 0.2,
    arrow = arrow(
      length = unit(5, "pt"),
      type = "closed"
    )
  ) + 
  geom_nodes(
    color = "gray", 
    size = degree(g, mode = "out")/2
  ) +
  scale_color_identity() +
  theme_void() +
  theme(legend.position = "none")  +
  geom_nodelabel_repel(
    aes(label = name),
    box.padding = unit(0.5, "lines"),
    data = gg,
    size = 3,
    segment.colour = "black",
    colour = "white",
    fill = "grey36"
  ) +
  coord_cartesian(xlim = c(-0.2, 1.2), ylim = c(-0.2, 1.2))
```

## One-way, two-way, and three-way interactions

Finally, we can quantify the overall importance of interactions in different taxa models, as well as the importance of one- and two-way interactions for a specific taxon using the same bootstrapping approach. To quantify interaction importance, we use H-statistics from the [`hstats`](https://github.com/mayer79/hstats) package.

```{r interactions, fig.height=10}
int_ <- mrInteractions(
  yhats_rf,
  num_bootstrap = 100,
  feature = "Plas",
  top_int = 10
)

int_[[1]] / # overall plot
int_[[2]] / # individual plot for feature
int_[[3]]   # two way plot for feature
```

The first plot shows that interactions account for an average of 60% (with a 90% bootstrap uncertainty interval of 44–81%) of variation in predictions for *H.killangoi*, and less for the other taxa. The second plot shows that interactions involving host abundance most affect predictions of *Plasmodium* (although *H.zosteropis* is also important and, according to the uncertainty intervals, could be even more important). This trend is similar community-wide. The last plot shows that the interaction between *Haemoproteus* presence and host abundance is the most important two-way interaction for *Plasmodium*, but this isn’t true community-wide, as host abundance and *H.zosteropis* form the strongest interaction overall. Taken together, we can see that interactions between taxa are mediated by host abundance.

Finally, we can explore specific interactions in more detail using 2D partial dependence plots implemented in the [`flashlight`](https://mayer79.github.io/flashlight/) package. Below, we plot how the two-way interaction between host abundance and *H.zosteropis* impacts the probability of detecting *Plasmodium*.

```{r two_way_int}
fl <- mrFlashlight(
  yhats_rf_downSamp,
  response = "single",
  index = 4 #index=4 selects Plas
)

fl %>%
  light_profile2d(
    c("scale.prop.zos", "Hzosteropis"),
    data
  ) %>%
  plot() +
  theme_bw()
```

So, if *H.zosteropsis* is not present and the relative abundance of *Zosterops* species is low, the probability of observing *Plasmodium* is high (>0.7).

### References

Cook et al., 2021: https://doi.org/10.18651/RWP2021-12
Kuhn (2014): https://doi.org/10.48550/arXiv.1405.6974
Fountain-Jones et al(2021): https://doi.org/10.1111/1755-0998.13495